### 2024-11-18 ms/ms.tex
overleaf

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
%
% This file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements, including customized APA reference formatting.
%
% You may copy this file and give it your
% article name, and enter your text.
%
% guidelines and troubleshooting are here:

%% To submit your paper:
\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
%%%%%%%
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% The trackchanges package adds five new LaTeX commands:
%
%  \note[editor]{The note}
%  \annote[editor]{Text to annotate}{The note}
%  \add[editor]{Text to add}
%  \remove[editor]{Text to remove}
%  \change[editor]{Text to remove}{Text to add}
%
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

\draftfalse

%% Enter journal name below.
%% Choose from this list of Journals:
%
% JGR: Atmospheres
% JGR: Biogeosciences
% JGR: Earth Surface
% JGR: Oceans
% JGR: Planets
% JGR: Solid Earth
% JGR: Space Physics
% Global Biogeochemical Cycles
% Geophysical Research Letters
% Paleoceanography and Paleoclimatology
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resources Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science
% Geohealth
%
% ie, \journalname{Water Resources Research}

\journalname{Geophysical Research Letters}

\begin{document}

\title{=enter title here=}

\authors{D. Kennedy\affil{1,2}, D. M. Lawrence\affil{1}, and I. R. Simpson\affil{1}}

\affiliation{1}{Climate and Global Dynamics Laboratory, NSF National Center for Atmospheric Research, Boulder, CO}
\affiliation{2}{Earth Research Institute, University of California Santa Barbara, Santa Barbara, CA}

\correspondingauthor{Daniel Kennedy}{djk2120@ucar.edu}

%  Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences

\begin{keypoints}
\item enter point 1 here
\item enter point 2 here
\item enter point 3 here
\end{keypoints}

\begin{abstract}
[ enter your Abstract here ]
\end{abstract}

\section*{Plain Language Summary}
Enter your Plain Language Summary here or delete this section.
Here are instructions on writing a Plain Language Summary: 
https://www.agu.org/Share-and-Advocate/Share/Community/Plain-language-summary

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  BODY TEXT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

\section{Methods}

\section{Results and Discussion}

\section{Conclusions}

%%

%  Numbered lines in equations:
%  To add line numbers to lines in equations,
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

%% Enter Figures and Tables near as possible to where they are first mentioned:
%
% DO NOT USE \psfrag or \subfigure commands.
%
% Figure captions go below the figure.
% Acronyms used in figure captions will be spelled out in the final, published version.

% Table titles go above tables;  other caption information
%  should be placed in last line of the table, using
% \multicolumn2l{$^a$ This is a table note.}
% NOTE that there is no difference between table caption and table heading in the final, published version
%
%----------------
% EXAMPLE FIGURES
%
% \begin{figure}
% \includegraphics{example.png}
% \caption{caption}
% \end{figure}
%
% Giving latex a width will help it to scale the figure properly. A simple trick is to use \textwidth. Try this if large figures run off the side of the page.
% \begin{figure}
% \noindent\includegraphics[width=\textwidth]{anothersample.png}
%\caption{caption}
%\label{pngfiguresample}
%\end{figure}
%
%
% If you get an error about an unknown bounding box, try specifying the width and height of the figure with the natwidth and natheight options. This is common when trying to add a PDF figure without pdflatex.
% \begin{figure}
% \noindent\includegraphics[natwidth=800px,natheight=600px]{samplefigure.pdf}
%\caption{caption}
%\label{pdffiguresample}
%\end{figure}
%
%
% PDFLatex does not seem to be able to process EPS figures. You may want to try the epstopdf package.
%

%
% ---------------
% EXAMPLE TABLE
%
% \begin{table}
% \caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
% \centering
% \begin{tabular}{l c}
% \hline
%  Run  & Time (min)  \\
% \hline
%   $l1$  & 260   \\
%   $l2$  & 300   \\
%   $l3$  & 340   \\
%   $h1$  & 270   \\
%   $h2$  & 250   \\
%   $h3$  & 380   \\
%   $r1$  & 370   \\
%   $r2$  & 390   \\
% \hline
% \multicolumn{2}{l}{$^{a}$Footnote text here.}
% \end{tabular}
% \end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% SIDEWAYS FIGURES and TABLES
% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
%
% \begin{sidewaysfigure}
% \includegraphics[width=20pc]{figsamp}
% \caption{caption here}
% \label{newfig}
% \end{sidewaysfigure}
%
%  \begin{sidewaystable}
%  \caption{Caption here}
% \label{tab:signif_gap_clos}
%  \begin{tabular}{ccc}
% one&two&three\\
% four&five&six
%  \end{tabular}
%  \end{sidewaystable}

%% If using numbered lines, please surround equations with \begin{linenomath*}...\end{linenomath*}
%\begin{linenomath*}
%\begin{equation}
%y|{f} \sim g(m, \sigma),
%\end{equation}
%\end{linenomath*}

%%% End of body of article

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional Appendices go here
%
% The \appendix command resets counters and redefines section heads
%
% After typing \appendix
%
%\section{Here Is Appendix Title}
% will show
% A: Here Is Appendix Title
%
%\appendix
%\section{Here is a sample appendix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Optional Glossary, Notation or Acronym section goes here:
%
% Glossary is only allowed in Reviews of Geophysics
%  \begin{glossary}
%  \term{Term}
%   Term Definition here
%  \term{Term}
%   Term Definition here
%  \term{Term}
%   Term Definition here
%  \end{glossary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acronyms
%% NOTE that acronyms in the final published version will be spelled out when used in figure captions.
%   \begin{acronyms}
%   \acro{Acronym}
%   Definition here
%   \acro{EMOS}
%   Ensemble model output statistics
%   \acro{ECMWF}
%   Centre for Medium-Range Weather Forecasts
%   \end{acronyms}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Notation
%   \begin{notation}
%   \notation{$a+b$} Notation Definition here
%   \notation{$e=mc^2$}
%   Equation in German-born physicist Albert Einstein's theory of special
%  relativity that showed that the increased relativistic mass ($m$) of a
%  body comes from the energy of motion of the body—that is, its kinetic
%  energy ($E$)—divided by the speed of light squared ($c^2$).
%   \end{notation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% DATA SECTION and ACKNOWLEDGMENTS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Open Research Section}
This section MUST contain a statement that describes where the data supporting the conclusions can be obtained. Data cannot be listed as ''Available from authors'' or stored solely in supporting information. Citations to archived data should be included in your reference list. Wiley will publish it as a separate section on the paper’s page. Examples and complete information are here:
https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors

\acknowledgments
Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% REFERENCES and BIBLIOGRAPHY
%
% \bibliography{<name of your .bib file>} don't specify the file extension
% don't specify bibliographystyle
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\bibliography{ enter your bibtex bibliography filename here }

%Reference citation instructions and examples:
%
% Please use ONLY \cite and \citeA for reference citations.
% \cite for parenthetical references
% ...as shown in recent studies (Simpson et al., 2019)
% \citeA for in-text citations
% ...Simpson et al. (2019) have shown...
%
%
%...as shown by \citeA{jskilby}.
%...as shown by \citeA{lewin76}, \citeA{carson86}, \citeA{bartoldy02}, and \citeA{rinaldi03}.
%...has been shown \cite{jskilbye}.
%...has been shown \cite{lewin76,carson86,bartoldy02,rinaldi03}.
%... \cite <i.e.>[]{lewin76,carson86,bartoldy02,rinaldi03}.
%...has been shown by \cite <e.g.,>[and others]{lewin76}.
%
% apacite uses < > for prenotes and [ ] for postnotes
% DO NOT use other cite commands (e.g., \citet, \citep, \citeyear, \nocite, \citealp, etc.).
%

\end{document}

More Information and Advice:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  SECTION HEADS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Capitalize the first letter of each word (except for
% prepositions, conjunctions, and articles that are
% three or fewer letters).

% AGU follows standard outline style; therefore, there cannot be a section 1 without
% a section 2, or a section 2.3.1 without a section 2.3.2.
% Please make sure your section numbers are balanced.
% ---------------
% Level 1 head
%
% Use the \section{} command to identify level 1 heads;
% type the appropriate head wording between the curly
% brackets, as shown below.
%
%An example:
%\section{Level 1 Head: Introduction}
%
% ---------------
% Level 2 head
%
% Use the \subsection{} command to identify level 2 heads.
%An example:
%\subsection{Level 2 Head}
%
% ---------------
% Level 3 head
%
% Use the \subsubsection{} command to identify level 3 heads
%An example:
%\subsubsection{Level 3 Head}
%
%---------------
% Level 4 head
%
% Use the \subsubsubsection{} command to identify level 3 heads
% An example:
%\subsubsubsection{Level 4 Head} An example.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  IN-TEXT LISTS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Do not use bulleted lists; enumerated lists are okay.
% \begin{enumerate}
% \item
% \item
% \item
% \end{enumerate}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  EQUATIONS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Single-line equations are centered.
% Equation arrays will appear left-aligned.

Math coded inside display math mode \[ ...\]
 will not be numbered, e.g.,:
 \[ x^2=y^2 + z^2\]

 Math coded inside \begin{equation} and \end{equation} will
 be automatically numbered, e.g.,:
 \begin{equation}
 x^2=y^2 + z^2
 \end{equation}

% To create multiline equations, use the
% \begin{eqnarray} and \end{eqnarray} environment
% as demonstrated below.
\begin{eqnarray}
  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
        && + (y - y_{0}) \sin \Theta  \nonumber \\
  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
        && + (y - y_{0}) \cos \Theta.
\end{eqnarray}

%If you don't want an equation number, use the star form:
%\begin{eqnarray*}...\end{eqnarray*}

% Break each line at a sign of operation
% (+, -, etc.) if possible, with the sign of operation
% on the new line.

% Indent second and subsequent lines to align with
% the first character following the equal sign on the
% first line.

% Use an \hspace{} command to insert horizontal space
% into your equation if necessary. Place an appropriate
% unit of measure between the curly braces, e.g.
% \hspace{1in}; you may have to experiment to achieve
% the correct amount of space.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  EQUATION NUMBERING: COUNTER
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% You may change equation numbering by resetting
% the equation counter or by explicitly numbering
% an equation.

% To explicitly number an equation, type \eqnum{}
% (with the desired number between the brackets)
% after the \begin{equation} or \begin{eqnarray}
% command.  The \eqnum{} command will affect only
% the equation it appears with; LaTeX will number
% any equations appearing later in the manuscript
% according to the equation counter.
%

% If you have a multiline equation that needs only
% one equation number, use a \nonumber command in
% front of the double backslashes (\\) as shown in
% the multiline equation above.

% If you are using line numbers, remember to surround
% equations with \begin{linenomath*}...\end{linenomath*}

%  To add line numbers to lines in equations:
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

### 2024-11-18 ms/ms.tex
intro

@@
 \section{Introduction}
+
+The 2020 Southwest U.S. drought was the driest summer of the instrumented record and part of a 20-year megadrought with severe consequences for water resources.
+
+Coupled ESMs have not captured humidity trends in semi-arid ecosystems \cite{simpson2024}.
+
+Though ESMs show a slight increase in spring precipitation in the region, there does not seem to be an attributable influence of climate change on NAM precipitation.
+Previous work indicates that the 2020 SWUS drought was not exacerbated by climate change, but was rather a result of  internal atmospheric variability \cite{seager2022}.
 

### 2025-05-19 ms/ms.tex

@@
 
-%% Enter journal name below.
-%% Choose from this list of Journals:
-%
-% JGR: Atmospheres
-% JGR: Biogeosciences
-% JGR: Earth Surface
-% JGR: Oceans
-% JGR: Planets
-% JGR: Solid Earth
-% JGR: Space Physics
-% Global Biogeochemical Cycles
-% Geophysical Research Letters
-% Paleoceanography and Paleoclimatology
-% Radio Science
-% Reviews of Geophysics
-% Tectonics
-% Space Weather
-% Water Resources Research
-% Geochemistry, Geophysics, Geosystems
-% Journal of Advances in Modeling Earth Systems (JAMES)
-% Earth's Future
-% Earth and Space Science
-% Geohealth
-%
-% ie, \journalname{Water Resources Research}
-
 \journalname{Geophysical Research Letters}
@@
 \begin{abstract}
-[ enter your Abstract here ]
+The 2020 Southwest U.S.
+drought was the driest summer of the instrumented record and part of a 20-year megadrought with severe consequences for water resources.
+Despite biases in the fully coupled CESM2, we were able to reproduce observed precipitation in a constrained circulation framework, by nudging the large scale circulation to reanalysis winds.
+Previous work has attributed the precipitation anomalies primarily to natural variability as opposed to radiative forcing or warming.
+We ported the 2020 circulation pattern to past and future climates to understand if warming influences monsoon period precipitation.
+Ensemble mean precipitation was 4mm higher under pre-industrial conditions, which is relatively small compared to the 2020 drought anomaly of -66mm, confirming that warming was at most a small contributor to monsoon failure.
+Ensemble mean precipitation was another 9mm lower under end-of-century conditions, indicating that future warming could further exacerbate Southwest U.S.
+droughts.
+However, we found that our future conditions constrained circulation ensemble was significantly drier than any simulations in a fully coupled framework, indicating that large-scale feedbacks may disallow this circulation pattern in the future.
 \end{abstract}
@@
 
-The 2020 Southwest U.S. drought was the driest summer of the instrumented record and part of a 20-year megadrought with severe consequences for water resources.
-
 Coupled ESMs have not captured humidity trends in semi-arid ecosystems \cite{simpson2024}.
@@
 \section{Results and Discussion}
+\subsection{Simulating the 2020 SWUS drought}
+
+\subsection{Porting the 2020 circulation to past and future climates}
+
+\subsection{Similarities and differences between the CCE and the CESM2-LE}
+ET$\sim$SM looks the same.
+But P$\sim$ET looks a bit different.
+The CCE mean looks to be in a reasonable spot, roughly speaking, but the partial response of P to varying ET (via land initialization) looks much smaller than in the fully coupled model.
+This would seem to indicate that the coupling of SM->ET is agnostic to the broader circulation, whereas P is not simply determined by the local ET.
+In the coupled model we tend to see an 18mm reduction in P for a 10mm reduction in ET, but in the CCE it looks more like 5mm P per 10mm ET.
+In the coupled model the circulation associated with 140mm ET may be quite different from the circulation that results in 120mm ET, in addition to any differences in the initial conditions.
+In the CCE, only the initial conditions are different, and the large-scale circulation is following observations.
+When CCE goes from ET of 120 to 100, the broader circulation still provides a relatively equivalent flux of water to the region (in fact slightly higher), which helps support the precipitation from falling drastically.
+In the coupled model, when ET goes from 140 to 120mm, it is probably largely driven by a change to a more diverging circulation, whereby the precip is robbed of the convergence flux as well as the associated reductions in ET flux.
+
+Does this invalidate inference from the CCE?
+No, it just means that any takeaways are conditional to the given circulation pattern.
+In the case of understanding sensitivity to initial conditions in the 2020 ensemble, the experimental framework is probably quite helpful.
+In the case of the analogue simulations, it's slightly more complicated, because warmer or colder climates may not be able to develop this circulation pattern.
+Towards this end, the fact that the 2090 ensemble is so much drier than anything seen in the fully coupled model is hard to overlook.
+This would seem to indicate that CESM2 predicts changes in circulation that would disallow the droughts we are simulating in the 2090 CCE.
+Further work to understand the mechanisms underpinning may be interesting, especially given the tendency across many coupled models to produce aberrant VP trends in semi-arid ecosystems.
+
+Why is the CCE drier than the LE in 2090?
+
+\subsection{Inference capabilities for extreme droughts}
+
+\begin{figure}
+\includegraphics[width=30pc]{../figs/main/precip.pdf}
+\caption{The constrained circulation ensemble provides a reasonable match for precipitation observations during the 2020 SWUS drought, despite  biases in the fully coupled model.
+Shaded areas span the 5th to 95th percentiles across pooled ensemble-years (CESM2: 1980-2020, CCE: 2020 only), or in the case of the GPCP observational product, simply across years (1980-2020).
+Solid lines denote the climatological averages.
+The GPCP observed precipitation during 2020 is shown as the dotted line.
+Additional panel with experimental protocol and perhaps some dynamical notion tk.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=40pc]{../figs/main/scatter_ET_P.png}
+\caption{Monsoon period precipitation vs.
+ET in the CCE and a dry subset of the CESM2-LE from three different time periods.
+ET and Precip seem to decline in both ensembles in response to climate change forcing.
+Printed numbers indicate the various ensemble means.
+Two asterisks indicate a given ensemble mean is distinguishable from both the other time period means at the p$<$0.05 level.
+One asterisk indicates it is distinguishable from just one of the other time periods.
+The dry subset of the CESM2-LE is defined as the 3\% driest ensemble years from the given time period based on JAS 10cm soil moisture (see methods).
+Relabeled as 1850 tk.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=35pc]{../figs/main/contours_ET_SW.pdf}
+\caption{Soil moisture (10cm) vs.
+ET in the CCE (dots) and the CESM2-LE (scatter density heat map) for three time periods.
+The CCE appears fully consistent with the coupled model with regard to this relationship, demonstrating an intensification of the hydrological cycle in the region.
+Note that the CCE does however appear significantly drier relative to the LE in 2090 as compared to 1850 or 2020.
+A second set of plots with something less clean looking potentially tk.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=35pc]{../figs/main/contours_PREC_ET.pdf}
+\caption{Soil moisture (10cm) vs.
+ET in the CCE (dots) and the CESM2-LE (scatter density heat map) for three time periods.
+The CCE appears fully consistent with the coupled model with regard to this relationship, demonstrating an intensification of the hydrological cycle in the region.
+Note that the CCE does however appear significantly drier relative to the LE in 2090 as compared to 1850 or 2020.
+A second set of plots with something less clean looking potentially tk.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=40pc]{../figs/main/cesm2le_water_anomalies.png}
+\caption{Monsoon period precipitation and ET anomalies (relative to 1970-1989) from the CESM2-LE from various ensemble subsets delineated based on 10cm soil moisture.
+Whereas the ensemble mean (turquoise) shows increases in precipitation and ET, the drier subsets can feature decreases.
+For reference, CESM2 climatological precipitation and ET are XX and YYmm.
+Refigured with non-overlapping composites tk.}
+\end{figure}
 
@@
 
-%\bibliography{ enter your bibtex bibliography filename here }
+\bibliography{refs.bib}
 

### 2025-05-27 ms/ms.tex
Updates from Overleaf

--- 2024-11-18 ms/ms.tex
@@
 
-Though ESMs show a slight increase in spring precipitation in the region, there does not seem to be an attributable influence of climate change on NAM precipitation.
+Though ESMs show a potential reduction in spring precipitation in the region, there does not seem to be an attributable influence of climate change on NAM precipitation.
 Previous work indicates that the 2020 SWUS drought was not exacerbated by climate change, but was rather a result of  internal atmospheric variability \cite{seager2022}.
@@
 
-%\bibliography{ enter your bibtex bibliography filename here }
+\bibliography{ms/refs}
 

### 2025-09-24 ms/cce.tex
ms

--- 2025-05-27 ms/ms.tex
@@
 \draftfalse
-
-%% Enter journal name below.
-%% Choose from this list of Journals:
-%
-% JGR: Atmospheres
-% JGR: Biogeosciences
-% JGR: Earth Surface
-% JGR: Oceans
-% JGR: Planets
-% JGR: Solid Earth
-% JGR: Space Physics
-% Global Biogeochemical Cycles
-% Geophysical Research Letters
-% Paleoceanography and Paleoclimatology
-% Radio Science
-% Reviews of Geophysics
-% Tectonics
-% Space Weather
-% Water Resources Research
-% Geochemistry, Geophysics, Geosystems
-% Journal of Advances in Modeling Earth Systems (JAMES)
-% Earth's Future
-% Earth and Space Science
-% Geohealth
-%
-% ie, \journalname{Water Resources Research}
-
 \journalname{Geophysical Research Letters}
@@
 
-\title{=enter title here=}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%  TITLE
+%
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\authors{D. Kennedy\affil{1,2}, D. M. Lawrence\affil{1}, and I. R. Simpson\affil{1}}
+\title{Constrained Circulation Ensemble}
 
-\affiliation{1}{Climate and Global Dynamics Laboratory, NSF National Center for Atmospheric Research, Boulder, CO}
-\affiliation{2}{Earth Research Institute, University of California Santa Barbara, Santa Barbara, CA}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  AUTHORS AND AFFILIATIONS
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+\authors{D. Kennedy\affil{1,2}, D. M. Lawrence\affil{1}, I. R. Simpson\affil{1}}
+\affiliation{1}{Climate and Global Dynamics Laboratory, NSF-NCAR}
+\affiliation{2}{Earth Research Institute, University of California Santa Barbara}
 \correspondingauthor{Daniel Kennedy}{djk2120@ucar.edu}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% KEY POINTS
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%  List up to three key points (at least one is required)
+%  Key Points summarize the main points and conclusions of the article
 %  Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
+
+% Example:
+% \begin{keypoints}
+% \item	List up to three key points (at least one is required)
+% \item	Key Points summarize the main points and conclusions of the article
+% \item	Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
+% \end{keypoints}
 
@@
 \end{keypoints}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  ABSTRACT and PLAIN LANGUAGE SUMMARY
+%
+% A good Abstract will begin with a short description of the problem
+% being addressed, briefly describe the new data or analyses, then
+% briefly states the main conclusion(s) and how they are supported and
+% uncertainties.
+
+% The Plain Language Summary should be written for a broad audience,
+% including journalists and the science-interested public, that will not have 
+% a background in your field.
+%
+% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
+% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
+% see http://sharingscience.agu.org/creating-plain-language-summary/)
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% \begin{abstract} starts the second page
 
@@
 %
-%  BODY TEXT
+%  OUTLINE
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Outline and in flux}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  INTRODUCTION
 %
@@
 \section{Introduction}
+%Text here ===>>>
+Models incorrectly predict positive vapor pressure trends in semi-arid regions \cite{simpson2024}.
 
-The 2020 Southwest U.S. drought was the driest summer of the instrumented record and part of a 20-year megadrought with severe consequences for water resources.
-
-Coupled ESMs have not captured humidity trends in semi-arid ecosystems \cite{simpson2024}.
-
-Though ESMs show a potential reduction in spring precipitation in the region, there does not seem to be an attributable influence of climate change on NAM precipitation.
-Previous work indicates that the 2020 SWUS drought was not exacerbated by climate change, but was rather a result of  internal atmospheric variability \cite{seager2022}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  METHODS
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \section{Methods}
+%Text here ===>>>
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  DISCUSSION
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \section{Results and Discussion}
+%Text here ===>>>
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  CONCLUSION
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \section{Conclusions}
+%Text here ===>>>
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
 % REFERENCES and BIBLIOGRAPHY
-%
-% \bibliography{<name of your .bib file>} don't specify the file extension
-% don't specify bibliographystyle
 %
@@
 
-\bibliography{ms/refs}
+\bibliography{ refs.bib }
 

### 2025-09-24 ms/makebib.sh

@@ -1,4 +1,4 @@
-grep citation ms.aux | sed 's/citation//g' | tr -d '\\{}' | sort | uniq > cites.txt
+grep citation cce.aux | sed 's/citation//g' | tr -d '\\{}' | sort | uniq > cites.txt
 
 d='/Users/djk2120/projects/references/refs/'
 :> refs.bib

### 2025-10-09 ms/supp.tex
methods

\documentclass[11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[parfill]{parskip}
\usepackage[margin=1in]{geometry}
\usepackage{pdflscape}
\title{Supplementary information for ``Constrained circulation ensemble"}
\author{D. Kennedy}
\renewcommand{\thefigure}{S\arabic{figure}}
\begin{document}
\maketitle

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{../figs/tbd.png}
\caption{A plot showing monsoon period temperature and precip in the CCE, LENS2, and observations, including maps}
\label{supp:drought}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{../figs/tbd.png}
\caption{A plot to validate the study design, specifically zonal winds}
\label{supp:winds}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=40pc]{../figs/supp/scatter_ET_P_allpoints.pdf}
\caption{JJA precipitation vs. evapotranspiration, as in main text Figure 2, but instead of only showing the drought years, here we present all the data from the CESM2 large ensemble.}
\label{supp:scatter}
\end{figure}

\end{document}

### 2025-11-03 ms/cce.tex

@@
 %
-%  OUTLINE
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Outline and in flux}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
 %  INTRODUCTION
@@
 \section{Introduction}
-%Text here ===>>>
+The summer of 2020 was the driest of the meteorological record within the Southwest United States \cite{mankin2021,hoell2022}, capping off an extended twenty year megadrought \cite{williams2022}.
 Models incorrectly predict positive vapor pressure trends in semi-arid regions \cite{simpson2024}.
@@
 \section{Methods}
-%Text here ===>>>
+
+Our study focused on the `four corners' regions of the western United States, which is defined as the region encompassing four adjoining states: Colorado, Utah, New Mexico, and Arizona.
+This region was chosen due to the spatial footprint of the 2020 drought (Supp Figure 1), as well as its significance for water management \cite{mankin2021} 
+Our goal was to create a large sample of simulations exhibiting extreme drought conditions, while still allowing for local land-atmosphere coupling, in order to investigate the influences of antecedent conditions and climate change.
+
+We ran multiple ensembles of simulations using CESM2, each consisting of 30 members, with varied initial conditions. For each simulation, winds were nudged to the ERA5 reanalysis for the year 2020. 
+Nudging occurred four times per day in each grid cell outside the box bounded by 28$^{\circ}$N, 47$^{\circ}$N, 130$^{\circ}$W, 96$^{\circ}$W.
+This serves to constrain the large-scale circulation, imposing meteorological drought conditions for every ensemble member.
+Inside the bounding box, the atmosphere evolves freely, allowing for full land-atmosphere coupling, which is necessary for studying the effects of initial conditions.
+Sea surface temperatures (SSTs) were imposed from an ERA5 reanalysis product, and all other forcing (e.g.
+CO$_2$, aerosols) followed the standard conventions for transient simulations as in the CESM2-LE.
+The simulations were initialized in April, using random initial conditions from the CESM2-LE for all the requisite land and atmosphere states (e.g. temperature, soil moisture, humidity).
+
+Our primary ensemble was based in 2020, mirroring the extreme 2020 Southwest U.S.
+drought.
+We also simulated this drought during pre-industrial and future conditions to investigate the influence of climate change.
+All of the ensembles were nudged to the 2020 winds, but initial conditions and other climate forcings were drawn from the appropriate time period mirroring the CESM2-LE, effectively porting the 2020 circulation to these two alternative climates.
+SSTs were anomaly-based, computing anomalies from the CESM2-LE and applying them to the ERA5 reanalysis SSTs used in the 2020 ensemble.
+Initial conditions were drawn at random from the CESM2-LE from either April 1850 or April 2090, and all other forcing (e.g.
+CO$_2$, aerosols) followed the standard conventions for transient simulations as in the CESM2-LE, which follows the SSP3-7.0 emissions scenario.
+
+We have compared our ensembles to the existing CESM2-LE \cite{rodgers2021}, which consists of 100 transient simulations (1850-2100) of the fully coupled configuration of CESM2.
+This ensemble is useful, because it provides a large number of model realizations, sampling across a range of internal climate variability.
+For the sake of comparison, we have utilized the same model source code and utilized the same forcing where applicable.
+The two key differences between our ensembles and the CESM2-LE are that 1) we impose the 2020 drought circulation via nudging and 2) we forced the simulations with observed SSTs (in lieu of dynamically coupled SSTs).
+
+ \begin{table}[h]
+ \caption{Ensemble descriptions}
+ \centering
+ \begin{tabular}{l c c c c}
+ \hline
+  Name  & Winds & Climate forcing & SSTs & Initial Conditions \\
+ \hline
+   Control  & ERA5-2020& 2020 (SSP3-7.0) & ERA5-2020 & CESM2-LE (2020)\\
+   Pre-industrial  & ERA5-2020 & 1850 & ERA5-2020 + CESM2-LE anomalies (1850) & CESM2-LE (1850) \\
+   Future        & ERA5-2020 & 2090 (SSP3-7.0) & ERA5-2020 + CESM2-LE anomalies (2020) & CESM2-LE (2090) \\
+ \hline
+ \end{tabular}
+ \label{tab:exps}
+ \end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  OUTLINE
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Outline and in flux}
+
+\subsection{Key points}
+\begin{itemize}
+	\item CCE can reproduce the 2020 drought
+	\item Soil water initialization influences drought severity via variables (soil moisture, runoff, temperature extremes), but does not exacerbate the meteorological drought, itself 
+	\item Warming affects this drought
+	\begin{itemize}
+		\item ET reduced in both free-running and constrained droughts
+		\item P reduced in CCE, but mixed signal in free-running CESM2-LE
+	\end{itemize}
+	\item 2090 CCE is much drier than the CESM2-LE
+\end{itemize}
+
+\subsection{Some numbers I'd like}
+\begin{itemize}
+\item 2020 JAS Precip (obs or CCE?) sits at the XXth percentile of the CESM2 large ensemble. 
+\end{itemize}
+
+\subsection{Supp figures I'll need}
+\begin{itemize}
+\item CESM2-LE JAS Precip historgram with obs and/or CCE
+\end{itemize}
 
@@
 \section{Results and Discussion}
-%Text here ===>>>
+
+\begin{figure}
+\includegraphics[width=32pc]{../figs/main/domain.pdf}
+\caption{Map of the study domain and nudging boundary.
+All analyses were based in the `four corners' region of the western United States, yellow area.
+Model simulations were nudged to reanalysis winds outside of the red box, in order to induce the observed large-scale circulation, while allowing for local land-atmosphere coupling.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=28pc]{../figs/main/precip.pdf}
+\caption{Precipitation climatologies for the four corners region from reanalysis (GPCP, 1981-2020) and a model large ensemble (CESM2-LE, 1981-2020) alongside our constrained circulation ensemble (CCE-2020).
+For each of the datasets, shading spans the 5th to 95th percentiles (across years and/or ensemble members), and the solid line tracks the mean.
+The dotted black line shows the reanalysis precipitation for 2020.
+The CESM2-LE shows a small high bias in mean and variance relative to reanalysis precipitation, but does not effectively capture the timing or magnitude of the summer monsoon.
+When nudged to 2020 winds, CESM2 can reproduce the 2020 drought, even though it falls outside the large ensemble envelope.
+N.b. will combine this with Figure 1.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=35pc]{../figs/main/inits.pdf}
+\caption{The influence of soil water initialization on various summer (JAS) drought indicators in the CCE, shown in the context of the variance across the CESM2-LE.
+In each case, the 10cm soil water content at initialization is plotted (red dots) against the summer drought indicator, with linear regression where appropriate (red lines).
+The colored stripes span the CESM2-LE deciles for the given drought indicator, truncated at the 1st and 99th percentiles.
+Dots in the white background area, exist below the 1st percentile of the CESM2-LE.
+Initial soil water content influences many drought indicators, but not the severity of the meteorological drought itself, i.e.
+summer precipitation.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=35pc]{../figs/main/scatter_ET_P.png}
+\caption{Summer evapotranspiration vs.
+precipitation in the CCE (a) and a dry subset of the CESM2-LE (d), alongside histograms of ET (b,e) and precipitation (c,f) across three time periods.
+The CESM2-LE is subset to the 30 driest ensemble-years for each period from 1000 total ensemble-years (100 members, 10 years around the given year) based on JAS 10cm soil water content.
+Stars indicate significant differences in the mean relative to the two other time periods, via t-test at the p$<$0.05 level.
+Forced responses tend to reduce ET and precipitation in both the CCE and the dry subset of the CESM2-LE.
+}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=35pc]{../figs/main/contours_ET_SW.pdf}
+\caption{Summer evapotranspiration vs.
+10cm soil water content from the CCE (dots) and the CESM2-LE (scatter density contours) across the three time periods.
+The relationship in the CCE closely follows the CESM2-LE pattern.
+The CCE exists at the dry extreme of the CESM2-LE, or even beyond, in the 2090 ensemble.
+}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=35pc]{../figs/main/contours_PREC_ET.pdf}
+\caption{Summer precipitation vs.
+evapotranspiration from the CCE (dots) and the CESM2-LE (scatter density contours) across the three time periods.
+The location of the CCE data comports with the CESM2-LE pattern, but the slopes do not appear to match.
+N.b.
+will combine this with Figure 5.}
+\end{figure}
 
@@
 %Text here ===>>>
-
-%%
-
-%  Numbered lines in equations:
-%  To add line numbers to lines in equations,
-%  \begin{linenomath*}
-%  \begin{equation}
-%  \end{equation}
-%  \end{linenomath*}
-
-%% Enter Figures and Tables near as possible to where they are first mentioned:
-%
-% DO NOT USE \psfrag or \subfigure commands.
-%
-% Figure captions go below the figure.
-% Acronyms used in figure captions will be spelled out in the final, published version.
-
-% Table titles go above tables;  other caption information
-%  should be placed in last line of the table, using
-% \multicolumn2l{$^a$ This is a table note.}
-% NOTE that there is no difference between table caption and table heading in the final, published version
-%
-%----------------
-% EXAMPLE FIGURES
-%
-% \begin{figure}
-% \includegraphics{example.png}
-% \caption{caption}
-% \end{figure}
-%
-% Giving latex a width will help it to scale the figure properly. A simple trick is to use \textwidth. Try this if large figures run off the side of the page.
-% \begin{figure}
-% \noindent\includegraphics[width=\textwidth]{anothersample.png}
-%\caption{caption}
-%\label{pngfiguresample}
-%\end{figure}
-%
-%
-% If you get an error about an unknown bounding box, try specifying the width and height of the figure with the natwidth and natheight options.
-This is common when trying to add a PDF figure without pdflatex.
-% \begin{figure}
-% \noindent\includegraphics[natwidth=800px,natheight=600px]{samplefigure.pdf}
-%\caption{caption}
-%\label{pdffiguresample}
-%\end{figure}
-%
-%
-% PDFLatex does not seem to be able to process EPS figures. You may want to try the epstopdf package.
-%
-
-%
-% ---------------
-% EXAMPLE TABLE
-%
-% \begin{table}
-% \caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
-% \centering
-% \begin{tabular}{l c}
-% \hline
-%  Run  & Time (min)  \\
-% \hline
-%   $l1$  & 260   \\
-%   $l2$  & 300   \\
-%   $l3$  & 340   \\
-%   $h1$  & 270   \\
-%   $h2$  & 250   \\
-%   $h3$  & 380   \\
-%   $r1$  & 370   \\
-%   $r2$  & 390   \\
-% \hline
-% \multicolumn{2}{l}{$^{a}$Footnote text here.}
-% \end{tabular}
-% \end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% SIDEWAYS FIGURES and TABLES
-% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
-%
-% \begin{sidewaysfigure}
-% \includegraphics[width=20pc]{figsamp}
-% \caption{caption here}
-% \label{newfig}
-% \end{sidewaysfigure}
-%
-%  \begin{sidewaystable}
-%  \caption{Caption here}
-% \label{tab:signif_gap_clos}
-%  \begin{tabular}{ccc}
-% one&two&three\\
-% four&five&six
-%  \end{tabular}
-%  \end{sidewaystable}
-
-%% If using numbered lines, please surround equations with \begin{linenomath*}...\end{linenomath*}
-%\begin{linenomath*}
-%\begin{equation}
-%y|{f} \sim g(m, \sigma),
-%\end{equation}
-%\end{linenomath*}
-
-%%% End of body of article
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Optional Appendices go here
-%
-% The \appendix command resets counters and redefines section heads
-%
-% After typing \appendix
-%
-%\section{Here Is Appendix Title}
-% will show
-% A: Here Is Appendix Title
-%
-%\appendix
-%\section{Here is a sample appendix}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Optional Glossary, Notation or Acronym section goes here:
-%
-% Glossary is only allowed in Reviews of Geophysics
-%  \begin{glossary}
-%  \term{Term}
-%   Term Definition here
-%  \term{Term}
-%   Term Definition here
-%  \term{Term}
-%   Term Definition here
-%  \end{glossary}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Acronyms
-%% NOTE that acronyms in the final published version will be spelled out when used in figure captions.
-%   \begin{acronyms}
-%   \acro{Acronym}
-%   Definition here
-%   \acro{EMOS}
-%   Ensemble model output statistics
-%   \acro{ECMWF}
-%   Centre for Medium-Range Weather Forecasts
-%   \end{acronyms}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Notation
-%   \begin{notation}
-%   \notation{$a+b$} Notation Definition here
-%   \notation{$e=mc^2$}
-%   Equation in German-born physicist Albert Einstein's theory of special
-%  relativity that showed that the increased relativistic mass ($m$) of a
-%  body comes from the energy of motion of the body—that is, its kinetic
-%  energy ($E$)—divided by the speed of light squared ($c^2$).
-%   \end{notation}
 
@@
 
-%Reference citation instructions and examples:
-%
-% Please use ONLY \cite and \citeA for reference citations.
-% \cite for parenthetical references
-% ...as shown in recent studies (Simpson et al., 2019)
-% \citeA for in-text citations
-% ...Simpson et al. (2019) have shown...
-%
-%
-%...as shown by \citeA{jskilby}.
-%...as shown by \citeA{lewin76}, \citeA{carson86}, \citeA{bartoldy02}, and \citeA{rinaldi03}.
-%...has been shown \cite{jskilbye}.
-%...has been shown \cite{lewin76,carson86,bartoldy02,rinaldi03}.
-%... \cite <i.e.>[]{lewin76,carson86,bartoldy02,rinaldi03}.
-%...has been shown by \cite <e.g.,>[and others]{lewin76}.
-%
-% apacite uses < > for prenotes and [ ] for postnotes
-% DO NOT use other cite commands (e.g., \citet, \citep, \citeyear, \nocite, \citealp, etc.).
-%
-
 \end{document}
-
-More Information and Advice:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  SECTION HEADS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Capitalize the first letter of each word (except for
-% prepositions, conjunctions, and articles that are
-% three or fewer letters).
-
-% AGU follows standard outline style; therefore, there cannot be a section 1 without
-% a section 2, or a section 2.3.1 without a section 2.3.2.
-% Please make sure your section numbers are balanced.
-% ---------------
-% Level 1 head
-%
-% Use the \section{} command to identify level 1 heads;
-% type the appropriate head wording between the curly
-% brackets, as shown below.
-%
-%An example:
-%\section{Level 1 Head: Introduction}
-%
-% ---------------
-% Level 2 head
-%
-% Use the \subsection{} command to identify level 2 heads.
-%An example:
-%\subsection{Level 2 Head}
-%
-% ---------------
-% Level 3 head
-%
-% Use the \subsubsection{} command to identify level 3 heads
-%An example:
-%\subsubsection{Level 3 Head}
-%
-%---------------
-% Level 4 head
-%
-% Use the \subsubsubsection{} command to identify level 3 heads
-% An example:
-%\subsubsubsection{Level 4 Head} An example.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  IN-TEXT LISTS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Do not use bulleted lists; enumerated lists are okay.
-% \begin{enumerate}
-% \item
-% \item
-% \item
-% \end{enumerate}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  EQUATIONS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Single-line equations are centered.
-% Equation arrays will appear left-aligned.
-
-Math coded inside display math mode \[ ...\]
- will not be numbered, e.g.,:
- \[ x^2=y^2 + z^2\]
-
- Math coded inside \begin{equation} and \end{equation} will
- be automatically numbered, e.g.,:
- \begin{equation}
- x^2=y^2 + z^2
- \end{equation}
-
-% To create multiline equations, use the
-% \begin{eqnarray} and \end{eqnarray} environment
-% as demonstrated below.
-\begin{eqnarray}
-  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
-        && + (y - y_{0}) \sin \Theta  \nonumber \\
-  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
-        && + (y - y_{0}) \cos \Theta.
-\end{eqnarray}
-
-%If you don't want an equation number, use the star form:
-%\begin{eqnarray*}...\end{eqnarray*}
-
-% Break each line at a sign of operation
-% (+, -, etc.) if possible, with the sign of operation
-% on the new line.
-
-% Indent second and subsequent lines to align with
-% the first character following the equal sign on the
-% first line.
-
-% Use an \hspace{} command to insert horizontal space
-% into your equation if necessary. Place an appropriate
-% unit of measure between the curly braces, e.g.
-% \hspace{1in}; you may have to experiment to achieve
-% the correct amount of space.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  EQUATION NUMBERING: COUNTER
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% You may change equation numbering by resetting
-% the equation counter or by explicitly numbering
-% an equation.
-
-% To explicitly number an equation, type \eqnum{}
-% (with the desired number between the brackets)
-% after the \begin{equation} or \begin{eqnarray}
-% command.  The \eqnum{} command will affect only
-% the equation it appears with; LaTeX will number
-% any equations appearing later in the manuscript
-% according to the equation counter.
-%
-
-% If you have a multiline equation that needs only
-% one equation number, use a \nonumber command in
-% front of the double backslashes (\\) as shown in
-% the multiline equation above.
-
-% If you are using line numbers, remember to surround
-% equations with \begin{linenomath*}...\end{linenomath*}
-
-%  To add line numbers to lines in equations:
-%  \begin{linenomath*}
-%  \begin{equation}
-%  \end{equation}
-%  \end{linenomath*}

### 2025-11-05 ms/cce.tex
ms

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%  List up to three key points (at least one is required)
-%  Key Points summarize the main points and conclusions of the article
-%  Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
-
-% Example:
-% \begin{keypoints}
-% \item	List up to three key points (at least one is required)
-% \item	Key Points summarize the main points and conclusions of the article
-% \item	Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
-% \end{keypoints}
 
 \begin{keypoints}
-\item enter point 1 here
-\item enter point 2 here
-\item enter point 3 here
+	\item Our CCE can reproduce the 2020 SWUS drought
+	\item Soil water initialization influences drought severity (soil moisture, runoff, temperature extremes), but does not exacerbate the meteorological drought, itself 
+	\item Warming from climate change may have exacerbated this drought, but to a small extent
+	\item 2090 CCE is much drier than 2090 CESM2-LE
 \end{keypoints}
@@
 \section{Introduction}
-The summer of 2020 was the driest of the meteorological record within the Southwest United States \cite{mankin2021,hoell2022}, capping off an extended twenty year megadrought \cite{williams2022}.
-Models incorrectly predict positive vapor pressure trends in semi-arid regions \cite{simpson2024}.
+
+Water in the Southwest United States (SWUS) is somewhat scarce, but also has myriad productive uses including agriculture, public water supply, recreation, and supporting the local ecology.
+Likewise the abundance/scarcity of water is strongly related to fire risk and temperature extremes. 
+Climate change has added strain to the water supply, with drier air and reduced snowpack due to warming, as well as increased precipitation variability.
+The summer of 2020 was the driest of the meteorological record within the SWUS \cite{mankin2021,hoell2022} and arrived in the midst of a 22-year megadrought (2000-2021), which was drier than any other extended drought in a soil moisture reconstruction dating back to 800 C.E.
+\cite{williams2022}.
+One more sentence about the effects of the 2020 drought tk, e.g. water shortage and temperature extremes.
+
+Drought in the SWUS and its intersection with climate change, is challenging to understand from observations alone, and has been frequently investigated with the aid of global or regional climate models.
+While models tend to agree on trends in temperature, soil moisture, and runoff, precipitation is more difficult to predict across this domain, due to the complex orography and the dynamic nature of the North American Monsoon \cite{cook2020,pascale2017}.
+There is also evidence that models may underestimate the drying effects of climate change on near-surface atmospheric vapor pressure in semi-arid regions \cite{simpson2024}.
+The general consensus seems to be that climate change has indeed exacerbated the consequences of meteorological drought, but that the recent precipitation shortfalls themselves are primarily attributable to natural climate variability, including a phase change of the Pacific Decadal Oscillation.
+
+Climate model large ensembles are often used for discerning forced responses from natural climate variability. 
+The CESM2 large ensemble \cite{rodgers2021} provides 100 ensemble members with varied initial conditions, yielding 100 distinct realizations of the historical record, and a large sample size for investigating drought dynamics in the SWUS.
+Instead of one instance of the year 2020, there are 100, and when we expand to look at neighboring years, we can create large samples of SWUS summers, numbering in the thousands.
+However, the statistical basis can still become small when focusing explicitly on extreme quantiles, or when controlling for very specific conditions. 
+Attribution of the 2020 SWUS drought is challenging, partly because of our imperfect models, but also because we are trying to diagnose changes not in the mean behavior, but in the extreme quantiles.
+Climate hazards are generally sporadic, and the accompanying risks are often concentrated in the extreme quantiles of the associated probability distributions, e.g. ``the hundred-year flood".
+As such, to effectively characterize risk using climate models, especially given finite computing resources, it can be helpful to design experiments that can up-sample these hazard occurrences.
+
+\vspace{80px}
+
+\noindent \P4: ensemble boosting techniques
+	\begin{itemize}
+	\item AMIP simulations
+	\begin{itemize}
+		\item even with observed SSTs it can be difficult to reproduce droughts, due to the influence of atmospheric internal variability
+	\end{itemize}
+	\item branched fully coupled simulations
+	\item nudging
+	\end{itemize}
+
+\noindent \P5: our experiment and research questions
 
@@
 
-Our study focused on the `four corners' regions of the western United States, which is defined as the region encompassing four adjoining states: Colorado, Utah, New Mexico, and Arizona.
-This region was chosen due to the spatial footprint of the 2020 drought (Supp Figure 1), as well as its significance for water management \cite{mankin2021} 
-Our goal was to create a large sample of simulations exhibiting extreme drought conditions, while still allowing for local land-atmosphere coupling, in order to investigate the influences of antecedent conditions and climate change.
+\subsection{Model description}
+
+For this study we utilized the Community Earth System Model, version 2 (CESM2) \cite{danabasoglu2020}.
+We ran a suite of simulations in land-atmosphere coupled mode (namely the FHIST\_BGC component set) using a data ocean (simulations described in Section \ref{sect:exp}).
+We compared our simulations to the CESM2 large ensemble (CESM2-LE) \cite{rodgers2021}, which consists of 100 transient simulations (1850-2100) of the fully coupled configuration of CESM2.
+The simulations follow historical forcing through 2014 and then switch to the SSP3-7.0 emissions scenario.
+This ensemble is useful, because it provides a large number of model realizations, sampling across a range of internal climate variability states.
+For the sake of comparison, we have utilized the CESM2-LE codebase for our simulations and the same forcing where applicable.
+
+\subsection{Data}
+
+We opted to force our simulations with SSTs from ERA5, the fifth generation European Center for Medium Range Weather Forecasts atmospheric reanalysis product \cite{hersbach2020}.
+These data are available at 0.25$^{\circ}$ resolution, which we regridded to the CESM2 nominal 1$^{\circ}$ resolution.
+We likewise utilized the wind fields from ERA5, in order to nudge our simulations to the 2020 circulation (described further in Section \ref{sect:exp}).
+All other forcings (e.g. aerosols, CO$_2$, land use) follow the protocol of the CESM2-LE \cite{rodgers2021}.
+We compared our simulations to the Global Precipitation Climatology Project (GPCP) Monthly Analysis Product \cite{adler2003}.
+These data are available at 2.5$^{\circ}$ resolution, which we likewise regridded to the CESM2 nominal 1$^{\circ}$ resolution.
+
+\subsection{Experimental design}
+\label{sect:exp}
+
+Our study is focused on the `four corners' regions of the Southwest United States, which is defined as the region encompassing four adjoining states: Colorado, Utah, New Mexico, and Arizona.
+This region was chosen based on the spatial footprint of the 2020 drought (Supp Figure 1), as well as its significance for water management \cite{mankin2021}. 
+Our goal was to create a set of simulations with extreme drought conditions, while still allowing for local land-atmosphere coupling, in order to investigate the influences of antecedent conditions and climate change.
 
 We ran multiple ensembles of simulations using CESM2, each consisting of 30 members, with varied initial conditions. For each simulation, winds were nudged to the ERA5 reanalysis for the year 2020. 
-Nudging occurred four times per day in each grid cell outside the box bounded by 28$^{\circ}$N, 47$^{\circ}$N, 130$^{\circ}$W, 96$^{\circ}$W.
+Nudging occurred four times per day in each grid cell outside the box bounded by 28$^{\circ}$N, 47$^{\circ}$N, 130$^{\circ}$W, 96$^{\circ}$W (Figure \ref{fig:domain}).
 This serves to constrain the large-scale circulation, imposing meteorological drought conditions for every ensemble member.
-Inside the bounding box, the atmosphere evolves freely, allowing for full land-atmosphere coupling, which is necessary for studying the effects of initial conditions.
+Inside the bounding box, the atmosphere evolves freely, allowing for local land-atmosphere coupling, which is necessary for studying the effects of initial conditions.
 Sea surface temperatures (SSTs) were imposed from an ERA5 reanalysis product, and all other forcing (e.g.
@@
 The simulations were initialized in April, using random initial conditions from the CESM2-LE for all the requisite land and atmosphere states (e.g. temperature, soil moisture, humidity).
+Each simulation ran for nine months.
 
@@
 We also simulated this drought during pre-industrial and future conditions to investigate the influence of climate change.
-All of the ensembles were nudged to the 2020 winds, but initial conditions and other climate forcings were drawn from the appropriate time period mirroring the CESM2-LE, effectively porting the 2020 circulation to these two alternative climates.
-SSTs were anomaly-based, computing anomalies from the CESM2-LE and applying them to the ERA5 reanalysis SSTs used in the 2020 ensemble.
+All of the ensembles were nudged to the 2020 winds, but initial conditions and other climate forcings were drawn from the appropriate time period of the CESM2-LE, effectively porting the 2020 circulation to these two alternative climates.
+SST forcing was anomaly-based, computing anomalies from the CESM2-LE and applying them to the ERA5 reanalysis SSTs used in the 2020 ensemble.
 Initial conditions were drawn at random from the CESM2-LE from either April 1850 or April 2090, and all other forcing (e.g.
 CO$_2$, aerosols) followed the standard conventions for transient simulations as in the CESM2-LE, which follows the SSP3-7.0 emissions scenario.
-
-We have compared our ensembles to the existing CESM2-LE \cite{rodgers2021}, which consists of 100 transient simulations (1850-2100) of the fully coupled configuration of CESM2.
-This ensemble is useful, because it provides a large number of model realizations, sampling across a range of internal climate variability.
-For the sake of comparison, we have utilized the same model source code and utilized the same forcing where applicable.
 The two key differences between our ensembles and the CESM2-LE are that 1) we impose the 2020 drought circulation via nudging and 2) we forced the simulations with observed SSTs (in lieu of dynamically coupled SSTs).
@@
    Pre-industrial  & ERA5-2020 & 1850 & ERA5-2020 + CESM2-LE anomalies (1850) & CESM2-LE (1850) \\
-   Future        & ERA5-2020 & 2090 (SSP3-7.0) & ERA5-2020 + CESM2-LE anomalies (2020) & CESM2-LE (2090) \\
+   Future        & ERA5-2020 & 2090 (SSP3-7.0) & ERA5-2020 + CESM2-LE anomalies (2090) & CESM2-LE (2090) \\
  \hline
@@
 %
-%  OUTLINE
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Outline and in flux}
-
-\subsection{Key points}
-\begin{itemize}
-	\item CCE can reproduce the 2020 drought
-	\item Soil water initialization influences drought severity via variables (soil moisture, runoff, temperature extremes), but does not exacerbate the meteorological drought, itself 
-	\item Warming affects this drought
-	\begin{itemize}
-		\item ET reduced in both free-running and constrained droughts
-		\item P reduced in CCE, but mixed signal in free-running CESM2-LE
-	\end{itemize}
-	\item 2090 CCE is much drier than the CESM2-LE
-\end{itemize}
-
-\subsection{Some numbers I'd like}
-\begin{itemize}
-\item 2020 JAS Precip (obs or CCE?) sits at the XXth percentile of the CESM2 large ensemble. 
-\end{itemize}
-
-\subsection{Supp figures I'll need}
-\begin{itemize}
-\item CESM2-LE JAS Precip historgram with obs and/or CCE
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
 %  DISCUSSION
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
+\section{Results and Discussion}
 
-\section{Results and Discussion}
+One tk intro sentence to say, hello and welcome to the results section.
+The CESM2-LE slightly over-predicts precipitation mean and variance in the SWUS, compared to the GPCP reanalysis product across 1981-2020 (Figure \ref{fig:climo}).
+Furthermore, CESM2 seems to underestimate the relative contribution of summer rain to annual precipitation and does not accurately match the observed monsoon timing.
+Despite this, the regional monsoon period (JAS) precipitation statistics match very closely between the CESM2-LE and the GPCP reanalysis product (Supp Figure 3).
+The 2020 SWUS drought was very dry (JAS precip = 71mm, GPCP), which sits at the the 1.8th percentile of the CESM2 large ensemble, primarily driven by extremely dry August and September months.
+Despite the extreme nature of the 2020 monsoon failure, the CCE is able to reproduce the observed meteorological drought.
+
+The CCE provides 30 replicates of the drought-inducing circulation, initialized across a range of soil water conditions (see Section \ref{sect:exp} for details).
+Dry initialization exacerbates drought severity, with significant reductions in summer soil moisture, evapotranspiration (ET), runoff as well as increased temperature (Figure \ref{fig:inits}).
+Of the various drought indicators, ET was most closely correlated with soil water initialization (Figure \ref{fig:inits}c), but net biome production experienced the largest initialization effect (based on the number of contours crossed by the regression line, Figure \ref{fig:inits}e), where drier initial soils significantly degraded the carbon sink.
+Monsoon period precipitation itself was not correlated with soil water initialization, indicating that the meteorological drought severity does not appear sensitive to initial conditions, consistent with previous findings \cite{jiang2025}.
+One more tk sentence about how CCE is pretty nice for testing initial conditions.
+
+We likewise tested the CCE across three climate scenarios: pre-industrial, present-day, and late century in an SSP3-7.0 warming scenario.
+The CCE shows a climate change influence on the regional water cycle, where both ET and precipitation are reduced moving from the pre-industrial period to present-day, and again moving from present-day to the end of this century (Figure \ref{fig:scatter}a-c).
+CCE monsoon precipitation was reduced by 4mm (tk, precise this) on average to present-day, with another 9mm reduction expected with SSP3-7.0 by the end of the century.
+Given that the 2020 monsoon failure represented a 64.4mm shortfall relative to climatological precipitation (Supp Figure 3), the radiative forcing effect of -4mm is small relative to the overall drought signal, which seems primarily attributable to unforced climate variability \cite{seager2022}.
+As such, climate change may have exacerbated the monsoon failure, but only slightly, and could potentially increase the severity of future precipitation shortfalls by another 9mm or roughly 12\%. 
+However, this experiment cannot capture any coupled circulation feedbacks that could exacerbate or mitigate the radiative response. 
+A tk sentence here providing a literature synopsis of forced circulation changes.
+
+The CESM2-LE, which does include circulation feedbacks, likewise shows a drying response within the most droughted years, but with significantly more noise (Figure \ref{fig:scatter}d-f).
+Because the CCE constrains the large-scale circulation, there is much less variance within the three ensembles.
+In the CESM2-LE, the large-scale circulation evolves freely, yielding a unique circulation pattern for every ensemble member.
+The response of ET to forcing in the CESM2-LE is consistent between PI to present and present to future, both showing a downward trend. 
+Precipitation seems to decline slightly between present-day and PI, but appears equivalent when comparing present-day and future. 
+These results correspond to a subset of dry years in the CESM2-LE, where we are analyzing the 30 driest monsoons in each time period, based on 10cm soil water, because our interest is specifically in forced changes to extreme droughts.
+Importantly, the forced responses at the extreme quantiles are distinct from those at moderate or wet quantiles (Supp Figure 4), highlighting the importance of investigating extremes separately from mean changes.
+For example, the overall CESM2-LE response to climate change in the SWUS is higher evapotranspiration, but it occurs with a prominent dipole of dry years trending drier, and moderate and wet years trending towards more evaporation.
+(This dipole seems to be unique with CESM2, but I don't really want to introduce the other datasets at this point, unless we think it's especially important)
+
+There is a strong relationship between summer soil water and ET in both the CCE and the CESM2-LE (Figure \ref{fig:swcontours}).
+The CCE simulations vary only by the initial conditions, whereas the CESM2-LE data are sampled across varying initial conditions and circulations patterns (with strong covariance). 
+Despite this, the two ensembles show comparable ET$\sim$SW relationships, indicating that the ET flux is largely determined by the soil moisture state alone, and relatively agnostic to the specific circulation pattern.
+This is in contrast to the relationship between precipitation and ET, where the slopes of the two relationships look quite different (Figure \ref{fig:pcontours}).
+The CCE shows a much weaker relationship between P and ET than seen in the CESM2-LE, indicating that, constrained to the 2020 circulation pattern, adding more ET does not have a large effect on P in the model.
+In the CESM2-LE shifting from left to right in these plots is not only changing the soil water (as in the CCE), but is likewise traversing a range of circulation patterns from diverging to converging, with associated effects on both ET and P.
+The large discrepancy in the P$\sim$ET relationship between the CCE and the CESM2-LE seems to indicate that the circulation pattern has a stronger influence on precipitation than does the local water availability.
+As such, a CCE framework allows us to explore how the model would behave absent circulation feedbacks, and in turn diagnose potential circulation feedback results when comparing a CCE to a free-running large ensemble.
+
+In this case, we ported the 2020 drought circulation to PI and future climates, which imposes the climate change radiative forcings, without any coupled changes to the circulation pattern.
+In many ways the CCE behaves akin to the CESM2-LE, however the CCE appears much drier relative the the CESM2-LE in 2090 than it does in 1850 or 2020 (e.g.
+CCE centroid well outside the CESM2-LE contours in Figure \ref{fig:swcontours}).
+Whereas the consequences of the 2020 drought circulation are within the envelope of the CESM2-LE in the PI and 2020 experiments, porting the circulation to 2090 results in very high values of moisture flux divergence and extremely low surface humidity that exceed the most extreme droughts of the CESM2-LE (Supp Figure 5).
+This would seem to indicate that coupled circulation feedbacks partially mitigate extreme divergence events in CESM2.
+
+\noindent Things to potentially think about:
+\begin{itemize}
+\item Why is CCE-2090 so dry? (tk plot AMJ convergence histograms for supp, P-E for now, can complicate later)
+\end{itemize}
 
@@
 Model simulations were nudged to reanalysis winds outside of the red box, in order to induce the observed large-scale circulation, while allowing for local land-atmosphere coupling.}
+\label{fig:domain}
 \end{figure}
@@
 N.b. will combine this with Figure 1.}
+\label{fig:climo}
 \end{figure}
@@
 summer precipitation.}
+\label{fig:inits}
 \end{figure}
@@
 }
+\label{fig:scatter}
 \end{figure}
@@
 }
+\label{fig:swcontours}
 \end{figure}
@@
 will combine this with Figure 5.}
+\label{fig:pcontours}
 \end{figure}
@@
 \section{Conclusions}
+
+\begin{enumerate}
+\item Did climate change exacerbate the 2020 SWUS drought? Should we expect even more severe droughts with continued warming? 
+\item CCE can assert drought
+\item CCE excludes coupled circulation changes
+\item Ensemble boosting is useful and probably quite prudent
+\end{enumerate}
+
+As such, whereas there may have been a small contribution of climate change to the historic precipitation shortfall of 2020, it is not clear that future climate change would further exacerbate monsoon failure in the SWUS.
+
+The lack of coupled circulation changes in the CCE is what allows us to impose drought in all of our simulations, but it also means that a CCE will not behave identically to a fully coupled model.
 %Text here ===>>>
+
+\clearpage
 
@@
 \acknowledgments
-Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
+DK, IRS, and DML acknowledge funding from the National Oceanic and Atmospheric Administration Modeling, Analysis, Predictions and Projections program (award number NA20OAR4310413) and funding from the NSF National Center for Atmospheric Research, which is a major facility sponsored by the NSF under Cooperative Agreement No.
+1852977
+Global Precipitation Climatology Project (GPCP) Monthly Analysis Product data provided by the NOAA PSL, Boulder, Colorado, USA, from their website at https://psl.noaa.gov
 

### 2025-11-05 ms/supp.tex
ms

@@
 \centering
-\includegraphics[width=20pc]{../figs/tbd.png}
-\caption{A plot showing monsoon period temperature and precip in the CCE, LENS2, and observations, including maps}
-\label{supp:drought}
+\includegraphics[width=30pc]{../figs/supp/gpcp_prec_map.png}
+\caption{2020 GPCP JAS precipitation anomalies relative to the 1981-2020 climatology.}
 \end{figure}
@@
 \centering
-\includegraphics[width=20pc]{../figs/tbd.png}
-\caption{A plot to validate the study design, specifically zonal winds}
-\label{supp:winds}
+\includegraphics[width=30pc]{../figs/supp/cce_anom_map.png}
+\caption{CCE JAS precipitation anomalies relative to the CESM2-LE climatology.}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=20pc]{../figs/supp/prec_hists.png}
+\caption{JAS precipitation statistics are quite similar between the CESM2-LE and the GPCP reanalysis product. TK add the LENS2 mean, and should add `2020' to CCE}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/supp/p_and_et_quantiles.png}
+\caption{P and ET response to climate change varies by soil moisture quantile. E.g. dry quantiles see reduced ET in the future, and wet quantiles see enhanced ET. }
+\label{supp:quantiles}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=20pc]{../figs/hists/VP_hist.png}
+\caption{Show that 2090 CCE is super dry, plotting histograms of MFC and 2m VP.}
+\label{supp:dryness}
 \end{figure}

### 2025-11-06 ms/cce.tex
ms

@@
 
-One tk intro sentence to say, hello and welcome to the results section.
+One or two tk intro sentence to say, hello and welcome to the results section.
 The CESM2-LE slightly over-predicts precipitation mean and variance in the SWUS, compared to the GPCP reanalysis product across 1981-2020 (Figure \ref{fig:climo}).
@@
 However, this experiment cannot capture any coupled circulation feedbacks that could exacerbate or mitigate the radiative response. 
-A tk sentence here providing a literature synopsis of forced circulation changes.
+A tk sentence here providing a literature reference to forced circulation changes.
 
@@
 
+This study highlights the utility of a constrained circulation ensemble experiment for investigating climate extremes.
+
 \begin{enumerate}
-\item Did climate change exacerbate the 2020 SWUS drought? Should we expect even more severe droughts with continued warming? 
-\item CCE can assert drought
-\item CCE excludes coupled circulation changes
+\item basic summary of the paper
+\item Did climate change exacerbate the 2020 SWUS drought? Should we expect more severe droughts with continued warming? 
+\item Why is 2090 so dry?
 \item Ensemble boosting is useful and probably quite prudent

### 2025-11-10 ms/cce.tex
Updates from Overleaf

--- 2025-11-05 ms/cce.tex
@@
 \journalname{Geophysical Research Letters}
-
 \begin{document}
@@
 %
-%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \title{Constrained Circulation Ensemble}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
 %  AUTHORS AND AFFILIATIONS
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \authors{D. Kennedy\affil{1,2}, D. M. Lawrence\affil{1}, I. R. Simpson\affil{1}}
@@
 \begin{abstract}
-[ enter your Abstract here ]
+TKTK
 \end{abstract}
@@
 \section*{Plain Language Summary}
-Enter your Plain Language Summary here or delete this section.
-Here are instructions on writing a Plain Language Summary: 
-https://www.agu.org/Share-and-Advocate/Share/Community/Plain-language-summary
+TKTK
 
@@
 
-Water in the Southwest United States (SWUS) is somewhat scarce, but also has myriad productive uses including agriculture, public water supply, recreation, and supporting the local ecology.
-Likewise the abundance/scarcity of water is strongly related to fire risk and temperature extremes. 
+Water in the Southwest United States (SWUS) is scarce, but also has myriad productive uses including agriculture, public water supply, recreation, and supporting the local ecology.
+Likewise the abundance of water is strongly related to fire risk and temperature extremes. 
 Climate change has added strain to the water supply, with drier air and reduced snowpack due to warming, as well as increased precipitation variability.
@@
 This ensemble is useful, because it provides a large number of model realizations, sampling across a range of internal climate variability states.
-For the sake of comparison, we have utilized the CESM2-LE codebase for our simulations and the same forcing where applicable.
+For the sake of comparison, we utilized the CESM2-LE codebase for our simulations and the same forcing where applicable.
 

### 2025-11-10 ms/cce.tex
more intro

@@
 While models tend to agree on trends in temperature, soil moisture, and runoff, precipitation is more difficult to predict across this domain, due to the complex orography and the dynamic nature of the North American Monsoon \cite{cook2020,pascale2017}.
-There is also evidence that models may underestimate the drying effects of climate change on near-surface atmospheric vapor pressure in semi-arid regions \cite{simpson2024}.
-The general consensus seems to be that climate change has indeed exacerbated the consequences of meteorological drought, but that the recent precipitation shortfalls themselves are primarily attributable to natural climate variability, including a phase change of the Pacific Decadal Oscillation.
+There is also evidence that models may systematically underestimate the drying effects of climate change on near-surface atmospheric vapor pressure in semi-arid regions \cite{simpson2024}.
+As such, it can be difficult to make strong inference regarding precipitation variations in the region, using large-scale coupled climate models.
+The general consensus seems to be that climate change has indeed exacerbated drought in the SWUS in a general sense, but that the recent precipitation shortfalls themselves are primarily attributable to random (natural) atmospheric variations, with potential contributions from a shift towards a sustained negative phase of the Pacific Decadal Oscillation \cite{mankin2021,seager2022, jiang2025}.
 
-Climate model large ensembles are often used for discerning forced responses from natural climate variability. 
-The CESM2 large ensemble \cite{rodgers2021} provides 100 ensemble members with varied initial conditions, yielding 100 distinct realizations of the historical record, and a large sample size for investigating drought dynamics in the SWUS.
-Instead of one instance of the year 2020, there are 100, and when we expand to look at neighboring years, we can create large samples of SWUS summers, numbering in the thousands.
-However, the statistical basis can still become small when focusing explicitly on extreme quantiles, or when controlling for very specific conditions. 
-Attribution of the 2020 SWUS drought is challenging, partly because of our imperfect models, but also because we are trying to diagnose changes not in the mean behavior, but in the extreme quantiles.
-Climate hazards are generally sporadic, and the accompanying risks are often concentrated in the extreme quantiles of the associated probability distributions, e.g. ``the hundred-year flood".
-As such, to effectively characterize risk using climate models, especially given finite computing resources, it can be helpful to design experiments that can up-sample these hazard occurrences.
+Climate model large ensembles are often used for discerning forced responses from natural climate variability \cite{deser2020}. 
+The CESM2 large ensemble \cite{rodgers2021} provides 100 ensemble members with varied initial conditions, yielding 100 distinct realizations of the historical period, and a large sample size for investigating drought dynamics in the SWUS.
+Instead of one instance of the year 2020, there are 100, such that we can create large samples of SWUS summers, in order to better analyze drought dynamics.
+However, the statistical basis can still become small when focusing explicitly on extreme quantiles, or when controlling for very specific circulation criteria. 
+Attribution of the 2020 SWUS drought is challenging, partly because of our imperfect models, but also because we are trying to diagnose changes not in the mean behavior, but in the severity of rare events.
+Climate hazards are often sporadic, and the accompanying risks are typically concentrated in the extreme quantiles of the probability distributions, e.g. ``the hundred-year flood".
+As such, to effectively characterize risk using climate models, especially given finite computing resources, it can be helpful to design experiments that up-sample these hazard occurrences.
+
+One such technique is to run land-atmosphere coupled simulations forced with the SSTs associated with an extreme event. 
+It is typical to generate ensembles with various land/atmosphere initial conditions to quantify the range of outcomes expected from that specific SST evolution. 
+These simulations are cheaper and faster than fully coupled simulations, and can be very helpful for understanding the association of SST patterns with hazard risk \cite{hoell2022,seager2022}.
+However, if the SSTs are not directly implicated in drought risk, such ensembles will not meaningfully upsample drought events, as seems to be the case with the 2020 summer drought in the SWUS \cite{jiang2025}.
+Another technique, reinitialized ensemble boosting, involves branching a fully coupled simulation ahead of a key event, and introducing roundoff level atmospheric perturbations to generate an ensemble of simulations.
+This has been used to better understand the proximate drivers of heat extremes, and to demonstrate the existence of even more extreme heatwaves \cite{gessner2021, fischer2023}. 
+This technique is very effective at up-sampling the extreme event, but generally can only sample a small range of initial conditions.
 

### 2025-11-10 ms/cce.tex
intro

@@
 	\item Warming from climate change may have exacerbated this drought, but to a small extent
-	\item 2090 CCE is much drier than 2090 CESM2-LE
+	\item 2090 CCE is much drier than 2090 CESM2-LE, indicating potential drought mitigation from coupled circulation change
 \end{keypoints}
@@
 \cite{williams2022}.
-One more sentence about the effects of the 2020 drought tk, e.g. water shortage and temperature extremes.
+One more sentence about the effects of the 2020 drought TKTK, e.g. water shortage and temperature extremes.
 
@@
 However, if the SSTs are not directly implicated in drought risk, such ensembles will not meaningfully upsample drought events, as seems to be the case with the 2020 summer drought in the SWUS \cite{jiang2025}.
-Another technique, reinitialized ensemble boosting, involves branching a fully coupled simulation ahead of a key event, and introducing roundoff level atmospheric perturbations to generate an ensemble of simulations.
+Another technique, reinitialized ensemble boosting, involves branching a fully coupled simulation ahead of a key event, and introducing roundoff level atmospheric perturbations to generate an ensemble of short-duration, fully-coupled simulations.
 This has been used to better understand the proximate drivers of heat extremes, and to demonstrate the existence of even more extreme heatwaves \cite{gessner2021, fischer2023}. 
-This technique is very effective at up-sampling the extreme event, but generally can only sample a small range of initial conditions.
+This technique is very effective at up-sampling the extreme event, but will generally span only a small range of initial conditions.
 
-\vspace{80px}
-
-\noindent \P4: ensemble boosting techniques
-	\begin{itemize}
-	\item AMIP simulations
-	\begin{itemize}
-		\item even with observed SSTs it can be difficult to reproduce droughts, due to the influence of atmospheric internal variability
-	\end{itemize}
-	\item branched fully coupled simulations
-	\item nudging
-	\end{itemize}
-
-\noindent \P5: our experiment and research questions
+In this study we impose a constrained circulation ensemble approach, where we not only force with reanalysis SSTs, but also nudge winds to reanalysis from the 2020 SWUS drought. 
+This allows us to generate an ensemble with drought in every simulation, but that also spans a wide range of land initialization possibilities. 
+This approach was recently used to investigate the influence of soil moisture preconditioning on the 2021 Pacific Northwest heatwave \cite{duan2025}.
+Using this framework, along with an existing fully coupled large ensemble, we are able to ask:
+\begin{enumerate}
+\item Can a climate model reproduce the 2020 SWUS drought, forced with the large-scale circulation?
+\item How do antecedent conditions, specifically soil moisture, influence drought severity?
+\item To what extent is the 2020 SWUS drought attributable to unforced climate variability, and should we expect future climate change to exacerbate monsoon failure?
+\end{enumerate}
 

### 2025-11-14 ms/cce.tex
intersperse figs
combine figs

@@
 \begin{enumerate}
-\item Can a climate model reproduce the 2020 SWUS drought, forced with the large-scale circulation?
+\item Can a climate model reproduce the 2020 SWUS drought, when forced with the observed large-scale circulation?
 \item How do antecedent conditions, specifically soil moisture, influence drought severity?
@@
 
+\begin{figure}
+\includegraphics[width=32pc]{../figs/main/fig1.png}
+\caption{Map of the study domain and nudging boundary (a).
+All analyses were based in the `four corners' region of the western United States, yellow area.
+Model simulations were nudged to reanalysis winds outside of the red box, in order to induce the observed large-scale circulation, while allowing for local land-atmosphere coupling.
+Precipitation climatologies (b) for the four corners region from reanalysis (GPCP, 1981-2020) and a model large ensemble (CESM2-LE, 1981-2020), alongside our constrained circulation ensemble (CCE-2020).
+For each of the datasets, shading spans the 5th to 95th percentiles (across years and/or ensemble members), and the solid line tracks the mean.
+The dotted black line shows the reanalysis precipitation for 2020.
+The CESM2-LE shows a small high bias in mean and variance relative to reanalysis precipitation, and it does not accurately capture the timing or magnitude of the summer monsoon.
+When nudged to 2020 winds (i.e.
+CCE-2020), CESM2 can reproduce the 2020 drought, even though it falls outside the large ensemble envelope.}
+\label{fig:fig1}
+\end{figure}
+
 Our primary ensemble was based in 2020, mirroring the extreme 2020 Southwest U.S.
@@
 One or two tk intro sentence to say, hello and welcome to the results section.
-The CESM2-LE slightly over-predicts precipitation mean and variance in the SWUS, compared to the GPCP reanalysis product across 1981-2020 (Figure \ref{fig:climo}).
+The CESM2-LE slightly over-predicts precipitation mean and variance in the SWUS, compared to the GPCP reanalysis product across 1981-2020 (Figure \ref{fig:fig1}b).
 Furthermore, CESM2 seems to underestimate the relative contribution of summer rain to annual precipitation and does not accurately match the observed monsoon timing.
@@
 
-The CCE provides 30 replicates of the drought-inducing circulation, initialized across a range of soil water conditions (see Section \ref{sect:exp} for details).
-Dry initialization exacerbates drought severity, with significant reductions in summer soil moisture, evapotranspiration (ET), runoff as well as increased temperature (Figure \ref{fig:inits}).
-Of the various drought indicators, ET was most closely correlated with soil water initialization (Figure \ref{fig:inits}c), but net biome production experienced the largest initialization effect (based on the number of contours crossed by the regression line, Figure \ref{fig:inits}e), where drier initial soils significantly degraded the carbon sink.
-Monsoon period precipitation itself was not correlated with soil water initialization, indicating that the meteorological drought severity does not appear sensitive to initial conditions, consistent with previous findings \cite{jiang2025}.
-One more tk sentence about how CCE is pretty nice for testing initial conditions.
-
-We likewise tested the CCE across three climate scenarios: pre-industrial, present-day, and late century in an SSP3-7.0 warming scenario.
-The CCE shows a climate change influence on the regional water cycle, where both ET and precipitation are reduced moving from the pre-industrial period to present-day, and again moving from present-day to the end of this century (Figure \ref{fig:scatter}a-c).
-CCE monsoon precipitation was reduced by 4mm (tk, precise this) on average to present-day, with another 9mm reduction expected with SSP3-7.0 by the end of the century.
-Given that the 2020 monsoon failure represented a 64.4mm shortfall relative to climatological precipitation (Supp Figure 3), the radiative forcing effect of -4mm is small relative to the overall drought signal, which seems primarily attributable to unforced climate variability \cite{seager2022}.
-As such, climate change may have exacerbated the monsoon failure, but only slightly, and could potentially increase the severity of future precipitation shortfalls by another 9mm or roughly 12\%. 
-However, this experiment cannot capture any coupled circulation feedbacks that could exacerbate or mitigate the radiative response. 
-A tk sentence here providing a literature reference to forced circulation changes.
-
-The CESM2-LE, which does include circulation feedbacks, likewise shows a drying response within the most droughted years, but with significantly more noise (Figure \ref{fig:scatter}d-f).
-Because the CCE constrains the large-scale circulation, there is much less variance within the three ensembles.
-In the CESM2-LE, the large-scale circulation evolves freely, yielding a unique circulation pattern for every ensemble member.
-The response of ET to forcing in the CESM2-LE is consistent between PI to present and present to future, both showing a downward trend. 
-Precipitation seems to decline slightly between present-day and PI, but appears equivalent when comparing present-day and future. 
-These results correspond to a subset of dry years in the CESM2-LE, where we are analyzing the 30 driest monsoons in each time period, based on 10cm soil water, because our interest is specifically in forced changes to extreme droughts.
-Importantly, the forced responses at the extreme quantiles are distinct from those at moderate or wet quantiles (Supp Figure 4), highlighting the importance of investigating extremes separately from mean changes.
-For example, the overall CESM2-LE response to climate change in the SWUS is higher evapotranspiration, but it occurs with a prominent dipole of dry years trending drier, and moderate and wet years trending towards more evaporation.
-(This dipole seems to be unique with CESM2, but I don't really want to introduce the other datasets at this point, unless we think it's especially important)
-
-There is a strong relationship between summer soil water and ET in both the CCE and the CESM2-LE (Figure \ref{fig:swcontours}).
-The CCE simulations vary only by the initial conditions, whereas the CESM2-LE data are sampled across varying initial conditions and circulations patterns (with strong covariance). 
-Despite this, the two ensembles show comparable ET$\sim$SW relationships, indicating that the ET flux is largely determined by the soil moisture state alone, and relatively agnostic to the specific circulation pattern.
-This is in contrast to the relationship between precipitation and ET, where the slopes of the two relationships look quite different (Figure \ref{fig:pcontours}).
-The CCE shows a much weaker relationship between P and ET than seen in the CESM2-LE, indicating that, constrained to the 2020 circulation pattern, adding more ET does not have a large effect on P in the model.
-In the CESM2-LE shifting from left to right in these plots is not only changing the soil water (as in the CCE), but is likewise traversing a range of circulation patterns from diverging to converging, with associated effects on both ET and P.
-The large discrepancy in the P$\sim$ET relationship between the CCE and the CESM2-LE seems to indicate that the circulation pattern has a stronger influence on precipitation than does the local water availability.
-As such, a CCE framework allows us to explore how the model would behave absent circulation feedbacks, and in turn diagnose potential circulation feedback results when comparing a CCE to a free-running large ensemble.
-
-In this case, we ported the 2020 drought circulation to PI and future climates, which imposes the climate change radiative forcings, without any coupled changes to the circulation pattern.
-In many ways the CCE behaves akin to the CESM2-LE, however the CCE appears much drier relative the the CESM2-LE in 2090 than it does in 1850 or 2020 (e.g.
-CCE centroid well outside the CESM2-LE contours in Figure \ref{fig:swcontours}).
-Whereas the consequences of the 2020 drought circulation are within the envelope of the CESM2-LE in the PI and 2020 experiments, porting the circulation to 2090 results in very high values of moisture flux divergence and extremely low surface humidity that exceed the most extreme droughts of the CESM2-LE (Supp Figure 5).
-This would seem to indicate that coupled circulation feedbacks partially mitigate extreme divergence events in CESM2.
-
-\noindent Things to potentially think about:
-\begin{itemize}
-\item Why is CCE-2090 so dry? (tk plot AMJ convergence histograms for supp, P-E for now, can complicate later)
-\end{itemize}
-
-\begin{figure}
-\includegraphics[width=32pc]{../figs/main/domain.pdf}
-\caption{Map of the study domain and nudging boundary.
-All analyses were based in the `four corners' region of the western United States, yellow area.
-Model simulations were nudged to reanalysis winds outside of the red box, in order to induce the observed large-scale circulation, while allowing for local land-atmosphere coupling.}
-\label{fig:domain}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=28pc]{../figs/main/precip.pdf}
-\caption{Precipitation climatologies for the four corners region from reanalysis (GPCP, 1981-2020) and a model large ensemble (CESM2-LE, 1981-2020) alongside our constrained circulation ensemble (CCE-2020).
-For each of the datasets, shading spans the 5th to 95th percentiles (across years and/or ensemble members), and the solid line tracks the mean.
-The dotted black line shows the reanalysis precipitation for 2020.
-The CESM2-LE shows a small high bias in mean and variance relative to reanalysis precipitation, but does not effectively capture the timing or magnitude of the summer monsoon.
-When nudged to 2020 winds, CESM2 can reproduce the 2020 drought, even though it falls outside the large ensemble envelope.
-N.b. will combine this with Figure 1.}
-\label{fig:climo}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[t]
 \includegraphics[width=35pc]{../figs/main/inits.pdf}
@@
 \end{figure}
+
+The CCE provides 30 replicates of the drought-inducing circulation, initialized across a range of soil water conditions (see Section \ref{sect:exp} for details).
+Dry initialization exacerbates drought severity, with significant reductions in summer soil moisture, evapotranspiration (ET), runoff as well as increased temperature (Figure \ref{fig:inits}).
+Of the various drought indicators, ET was most closely correlated with soil water initialization (Figure \ref{fig:inits}c), but net biome production experienced the largest initialization effect (based on the number of contours crossed by the regression line, Figure \ref{fig:inits}e), where drier initial soils significantly degraded the carbon sink.
+Monsoon period precipitation itself was not correlated with soil water initialization, indicating that the meteorological drought severity does not appear sensitive to initial conditions, consistent with previous findings \cite{jiang2025}.
+One more tk sentence about how CCE is pretty nice for testing initial conditions.
 
@@
 
+We likewise tested the CCE across three climate scenarios: pre-industrial, present-day, and late century in an SSP3-7.0 warming scenario.
+The CCE shows a climate change influence on the regional water cycle, where both ET and precipitation are reduced moving from the pre-industrial period to present-day, and again moving from present-day to the end of this century (Figure \ref{fig:scatter}a-c).
+CCE monsoon precipitation was reduced by 4mm (tk, precise this) on average between the pre-industrial and present-day simulations, with another 9mm reduction expected with SSP3-7.0 by the end of the century.
+Given that the 2020 monsoon failure represented a 64.4mm shortfall relative to climatological precipitation (Supp Figure 3), the CCE scenario effect of -4mm is small relative to the overall drought signal.
+As such, the meteorological drought seems primarily attributable to unforced climate variability \cite{seager2022}.
+As such, climate change may have exacerbated the monsoon failure, but only slightly, and could potentially increase the severity of future precipitation shortfalls by another 9mm or roughly 12\%. 
+However, the CCE experiment cannot capture any coupled circulation feedbacks that could exacerbate or mitigate the radiative response. 
+A tk sentence here providing a literature reference to forced circulation changes.
+
+The CESM2-LE, which does include circulation feedbacks, likewise shows a drying response across the most drought years, but with significantly more noise (Figure \ref{fig:scatter}d-f).
+Because the CCE constrains the large-scale circulation, there is much less variance within each ensemble. 
+In the CESM2-LE, the large-scale circulation evolves freely, yielding a unique circulation pattern for every ensemble member.
+The response of ET to forcing in the CESM2-LE is consistent between PI to present and present to future, both showing a downward trend. 
+Precipitation seems to decline slightly from pre-industrial to present-day, but appears equivalent when comparing present-day and future. 
+These results correspond to a subset of dry years in the CESM2-LE, where we are analyzing the 30 driest monsoons in each time period (based on 10cm soil water), because our interest is specifically in forced changes to extreme droughts.
+Importantly, the forced responses at the extreme quantiles are distinct from those at moderate or wet quantiles (Supp Figure 4), highlighting the importance of investigating extremes separately from mean changes.
+For example, the overall CESM2-LE response to climate change in the SWUS is higher evapotranspiration, but it occurs with a prominent dipole of dry years trending drier, and moderate and wet years trending towards more evaporation.
+(This dipole seems to be unique with CESM2, but I don't really want to introduce the other datasets at this point, unless we think it's especially important)
+
+There is a strong relationship between summer soil water and ET in both the CCE and the CESM2-LE (Figure \ref{fig:swcontours}).
+The CCE simulations vary only by the initial conditions, whereas the CESM2-LE data are sampled across various initial conditions and circulations patterns (with strong covariance). 
+Despite this, the two ensembles show comparable ET$\sim$SW relationships, indicating that the ET flux is largely determined by the soil moisture state alone, and relatively agnostic to the specific circulation pattern.
+This is in contrast to the relationship between precipitation and ET, where the slopes of the two relationships look quite different (Figure \ref{fig:pcontours}).
+The CCE shows a much weaker relationship between P and ET than seen in the CESM2-LE, indicating that, constrained to the 2020 circulation pattern, adding more ET does not have a large effect on P in the model.
+In the CESM2-LE shifting from left to right in these plots is not only changing the soil water (as in the CCE), but is likewise traversing a range of circulation patterns from diverging to converging, with associated effects on both ET and P.
+The large discrepancy in the P$\sim$ET relationship between the CCE and the CESM2-LE seems to indicate that the circulation pattern has a stronger influence on precipitation than does the local water availability.
+As such, a CCE framework allows us to explore how the model would behave absent circulation feedbacks, and in turn diagnose potential circulation feedback results when comparing a CCE to a free-running large ensemble.
+
+In this case, we ported the 2020 drought circulation to PI and future climates, which imposes the climate change radiative forcings, without any coupled changes to the circulation pattern.
+In many ways the CCE behaves akin to the CESM2-LE, however the CCE appears much drier relative the the CESM2-LE in 2090 than it does in 1850 or 2020 (e.g.
+CCE centroid well outside the CESM2-LE contours in Figure \ref{fig:swcontours}).
+Whereas the consequences of the 2020 drought circulation are within the envelope of the CESM2-LE in the PI and 2020 experiments, porting the circulation to 2090 results in very high values of moisture flux divergence and extremely low surface humidity that exceed the most extreme droughts of the CESM2-LE (Supp Figure 5).
+This would seem to indicate that coupled circulation feedbacks may partially mitigate extreme divergence events in CESM2, under late-century SSP3-7.0 conditions.
+
+\noindent Things to potentially think about:
+\begin{itemize}
+\item Why is CCE-2090 so dry? (tk plot AMJ convergence histograms for supp, P-E for now, can complicate later)
+\end{itemize}
+
 \begin{figure}
@@
 The relationship in the CCE closely follows the CESM2-LE pattern.
-The CCE exists at the dry extreme of the CESM2-LE, or even beyond, in the 2090 ensemble.
+The CCE exists at the dry extreme of the CESM2-LE, or even beyond, as in the 2090 scenario.
 }
@@
 This study highlights the utility of a constrained circulation ensemble experiment for investigating climate extremes. 
+Our CCE is able to reproduce the 2020 SWUS drought, despite known biases in the fully coupled CESM2.
+This experiment demonstrates that antecedent soil moisture exacerbates drought severity, but not the precipitation amounts.
+We likewise found only a small influence of climate change on precipitation volume during the 2020 drought in the CCE.
+It showed a larger effect by the end of the 20th century, but subject to uncertainty due to potential circulation feedbacks.
 
-\begin{enumerate}
-\item basic summary of the paper
-\item Did climate change exacerbate the 2020 SWUS drought? Should we expect more severe droughts with continued warming? 
-\item Why is 2090 so dry?
-\item Ensemble boosting is useful and probably quite prudent
-\end{enumerate}
+The 2020 SWUS drought analogues herein impose several climate change effects (e.g. SST warming, radiative forcing, aerosol changes).
+However we are repeating the same circulation across all three time periods, such that if the 2020 drought circulation is rendered unattainable by some forced response, inference in the constrained system would not apply to the fully coupled system.
+We found some evidence of this in our experiments, where, for certain effects, the CCE showed stronger responses to climate change than the fully coupled CESM2-LE.
+As such, whereas there may have been a small contribution of climate change to the historic precipitation shortfall of 2020 ($\sim$6$\%$ in the CCE), it is not clear that future climate change would further exacerbate monsoon failure in the SWUS.
+While the 1850 and 2020 ensembles looked comparable, in 2090 the CCE features less precipitation, more moisture divergence, and drier near surface air than the driest corresponding years from the CESM2-LE.
+This may indicate a propensity for CESM2 to mitigate moisture flux divergence via circulation feedbacks, but would need to be investigated more specifically.
+This could be of particular interest, given that fully coupled earth system models show erroneous trends in near surface atmospheric vapor pressure \cite{simpson2024}.
 
-As such, whereas there may have been a small contribution of climate change to the historic precipitation shortfall of 2020, it is not clear that future climate change would further exacerbate monsoon failure in the SWUS.
+We have higher confidence in our inference with regard to investigating preconditioning effects using the CCE approach.
+Drier initializations were associated with lower ET, lower runoff, higher temperatures, and a weaker carbon sink, serving to significantly exacerbate the drought.
+However, initialization did not have a strong influence on precipitation, itself. 
+Precipitation in our analyses was generally controlled by the circulation pattern, more than by local water availability.
+The CCE approach is particularly useful for investigating the influence of antecedent conditions, because we are able to force drought in all simulations, but also sample a wide range of initial conditions
 
-The lack of coupled circulation changes in the CCE is what allows us to impose drought in all of our simulations, but it also means that a CCE will not behave identically to a fully coupled model.
-%Text here ===>>>
+Ensemble boosting is useful and probably quite prudent
 

### 2025-11-14 ms/cce.tex
edit

@@
 As such, whereas there may have been a small contribution of climate change to the historic precipitation shortfall of 2020 ($\sim$6$\%$ in the CCE), it is not clear that future climate change would further exacerbate monsoon failure in the SWUS.
-While the 1850 and 2020 ensembles looked comparable, in 2090 the CCE features less precipitation, more moisture divergence, and drier near surface air than the driest corresponding years from the CESM2-LE.
-This may indicate a propensity for CESM2 to mitigate moisture flux divergence via circulation feedbacks, but would need to be investigated more specifically.
-This could be of particular interest, given that fully coupled earth system models show erroneous trends in near surface atmospheric vapor pressure \cite{simpson2024}.
+While the 1850 and 2020 ensembles looked comparable, in 2090 the CCE is drier than almost every CESM2-LE simulation with regard to 2m vapor pressure and 10cm soil moisture.
+This may indicate a propensity for CESM2 to mitigate extreme moisture flux divergence via circulation feedbacks, but would need to be investigated more specifically.
+This could be of particular interest, given that fully coupled Earth system models show erroneous trends in near surface atmospheric vapor pressure \cite{simpson2024}.
 

### 2025-11-14 ms/cce.tex
conclusion

@@
 \begin{keypoints}
-	\item Our CCE can reproduce the 2020 SWUS drought
-	\item Soil water initialization influences drought severity (soil moisture, runoff, temperature extremes), but does not exacerbate the meteorological drought, itself 
-	\item Warming from climate change may have exacerbated this drought, but to a small extent
-	\item 2090 CCE is much drier than 2090 CESM2-LE, indicating potential drought mitigation from coupled circulation change
+	\item Our constrained circulation ensemble can reproduce the 2020 SWUS drought.
+	\item Soil water initialization influences drought severity (soil moisture, runoff, temperature extremes), but does not exacerbate the meteorological drought, itself.
+	\item Warming from climate change may have exacerbated the monsoon failure, but to a small extent.
+	\item Our 2020 constrained ensemble is comparable to droughts seen in a fully coupled large ensemble, and likewise for an 1850 analogue, but is drier than the fully coupled model in 2090.
 \end{keypoints}
@@
 We ran multiple ensembles of simulations using CESM2, each consisting of 30 members, with varied initial conditions. For each simulation, winds were nudged to the ERA5 reanalysis for the year 2020. 
-Nudging occurred four times per day in each grid cell outside the box bounded by 28$^{\circ}$N, 47$^{\circ}$N, 130$^{\circ}$W, 96$^{\circ}$W (Figure \ref{fig:domain}).
+Nudging occurred four times per day in each grid cell outside the box bounded by 28$^{\circ}$N, 47$^{\circ}$N, 130$^{\circ}$W, 96$^{\circ}$W (Figure \ref{fig:fig1}a).
 This serves to constrain the large-scale circulation, imposing meteorological drought conditions for every ensemble member.
@@
 
-\noindent Things to potentially think about:
-\begin{itemize}
-\item Why is CCE-2090 so dry? (tk plot AMJ convergence histograms for supp, P-E for now, can complicate later)
-\end{itemize}
-
 \begin{figure}
@@
 Precipitation in our analyses was generally controlled by the circulation pattern, more than by local water availability.
-The CCE approach is particularly useful for investigating the influence of antecedent conditions, because we are able to force drought in all simulations, but also sample a wide range of initial conditions
+The CCE approach is particularly useful for investigating the influence of antecedent conditions, because we are able to force the meteorological drought in all simulations, but also sample a wide range of initial conditions.
 
-Ensemble boosting is useful and probably quite prudent
+Because climate risk tends to be concentrated in extreme events, it is prudent to design experiments to study such events beyond their natural occurrence rate in free-running simulations.
+Likewise forced changes to the mean in any given variable may not transpose directly to the extreme quantiles. 
+For example, we found that for CESM2, climate change tends to increase ET in our study domain on average, but actually results in a decrease in ET during drought years. 
+Frameworks to better resolve extreme events include initial condition large ensembles, AMIP ensembles, branched reinitialization ensemble boosting, and the constrained circulation ensemble.
+We find the CCE framework to be a useful tool for investigating extreme events, particularly when the simulations can be analyzed relative to an unconstrained initial condition large ensemble.
 

### 2025-11-21 ms/cce.tex

@@
 \begin{keypoints}
-	\item Our constrained circulation ensemble can reproduce the 2020 SWUS drought.
+	\item A constrained circulation ensemble can reproduce the 2020 SWUS drought.
 	\item Soil water initialization influences drought severity (soil moisture, runoff, temperature extremes), but does not exacerbate the meteorological drought, itself.
 	\item Warming from climate change may have exacerbated the monsoon failure, but to a small extent.
-	\item Our 2020 constrained ensemble is comparable to droughts seen in a fully coupled large ensemble, and likewise for an 1850 analogue, but is drier than the fully coupled model in 2090.
+	\item The constrained 2020 drought and 1850 analog align with the distribution from the free running model, but the 2090 analog is drier.
 \end{keypoints}
@@
 
-Water in the Southwest United States (SWUS) is scarce, but also has myriad productive uses including agriculture, public water supply, recreation, and supporting the local ecology.
-Likewise the abundance of water is strongly related to fire risk and temperature extremes. 
+Water in the Southwest United States (SWUS) is scarce, but also has myriad important uses including agriculture, public water supply, recreation, and supporting the local ecology.
+The abundance of water is also strongly related to fire risk and temperature extremes. 
 Climate change has added strain to the water supply, with drier air and reduced snowpack due to warming, as well as increased precipitation variability.
@@
 There is also evidence that models may systematically underestimate the drying effects of climate change on near-surface atmospheric vapor pressure in semi-arid regions \cite{simpson2024}.
-As such, it can be difficult to make strong inference regarding precipitation variations in the region, using large-scale coupled climate models.
-The general consensus seems to be that climate change has indeed exacerbated drought in the SWUS in a general sense, but that the recent precipitation shortfalls themselves are primarily attributable to random (natural) atmospheric variations, with potential contributions from a shift towards a sustained negative phase of the Pacific Decadal Oscillation \cite{mankin2021,seager2022, jiang2025}.
+As such, it can be difficult to make strong inference regarding precipitation variations or trends in the region, using large-scale coupled climate models.
+The general consensus seems to be that climate change has indeed exacerbated drought in the SWUS in a general sense, but that the precipitation shortfalls of the extreme 2020 drought themselves are primarily attributable to random (natural) atmospheric variations, with potential contributions from a shift towards a sustained negative phase of the Pacific Decadal Oscillation \cite{mankin2021,seager2022, jiang2025}.
 
-Climate model large ensembles are often used for discerning forced responses from natural climate variability \cite{deser2020}. 
+Climate model large ensembles are often used to discern forced responses from natural climate variability \cite{deser2020}. 
 The CESM2 large ensemble \cite{rodgers2021} provides 100 ensemble members with varied initial conditions, yielding 100 distinct realizations of the historical period, and a large sample size for investigating drought dynamics in the SWUS.
-Instead of one instance of the year 2020, there are 100, such that we can create large samples of SWUS summers, in order to better analyze drought dynamics.
+Instead of one instance of the year 2020, there are 100, such that we can generate large samples of SWUS summers, in order to better analyze drought dynamics in the context of natural variability.
 However, the statistical basis can still become small when focusing explicitly on extreme quantiles, or when controlling for very specific circulation criteria. 
@@
 
-One such technique is to run land-atmosphere coupled simulations forced with the SSTs associated with an extreme event. 
-It is typical to generate ensembles with various land/atmosphere initial conditions to quantify the range of outcomes expected from that specific SST evolution. 
-These simulations are cheaper and faster than fully coupled simulations, and can be very helpful for understanding the association of SST patterns with hazard risk \cite{hoell2022,seager2022}.
+One such technique is to run land-atmosphere coupled simulations forced with observation-based SSTs from an extreme event to understand the association of SST patterns and antecedent conditions with hazard risk \cite{hoell2022,seager2022}.
 However, if the SSTs are not directly implicated in drought risk, such ensembles will not meaningfully upsample drought events, as seems to be the case with the 2020 summer drought in the SWUS \cite{jiang2025}.
-Another technique, reinitialized ensemble boosting, involves branching a fully coupled simulation ahead of a key event, and introducing roundoff level atmospheric perturbations to generate an ensemble of short-duration, fully-coupled simulations.
-This has been used to better understand the proximate drivers of heat extremes, and to demonstrate the existence of even more extreme heatwaves \cite{gessner2021, fischer2023}. 
+Another technique, reinitialized ensemble boosting, involves branching a fully coupled simulation ahead of a key model-based event, and introducing roundoff level atmospheric perturbations to generate an ensemble of short-duration, fully-coupled simulations.
+This has been used to better understand the proximate drivers of heat extremes, and to demonstrate the existence of even more extreme heatwaves than can be produced with a limited sample size \cite{gessner2021, fischer2023}.
 This technique is very effective at up-sampling the extreme event, but will generally span only a small range of initial conditions.
 
-In this study we impose a constrained circulation ensemble approach, where we not only force with reanalysis SSTs, but also nudge winds to reanalysis from the 2020 SWUS drought. 
+In this study we impose a constrained circulation ensemble approach, where we not only force with observation-based SSTs, but also nudge winds to reanalysis to capture the specific circulation patterns of the 2020 SWUS drought.
 This allows us to generate an ensemble with drought in every simulation, but that also spans a wide range of land initialization possibilities. 
-This approach was recently used to investigate the influence of soil moisture preconditioning on the 2021 Pacific Northwest heatwave \cite{duan2025}.
+This approach was recently effectively used to investigate the influence of soil moisture preconditioning on the 2021 Pacific Northwest heatwave \cite{duan2025}.
 Using this framework, along with an existing fully coupled large ensemble, we are able to ask:
 \begin{enumerate}
-\item Can a climate model reproduce the 2020 SWUS drought, when forced with the observed large-scale circulation?
+\item Can a climate model reproduce the 2020 SWUS drought when forced with the observed large-scale circulation?
 \item How do antecedent conditions, specifically soil moisture, influence drought severity?
-\item To what extent is the 2020 SWUS drought attributable to unforced climate variability, and should we expect future climate change to exacerbate monsoon failure?
+\item To what extent is the 2020 SWUS drought attributable to unforced climate variability, and should we expect future climate change to exacerbate the impacts of a drought-inducing circulation pattern like that which occurred in 2020?
 \end{enumerate}
@@
 We ran a suite of simulations in land-atmosphere coupled mode (namely the FHIST\_BGC component set) using a data ocean (simulations described in Section \ref{sect:exp}).
-We compared our simulations to the CESM2 large ensemble (CESM2-LE) \cite{rodgers2021}, which consists of 100 transient simulations (1850-2100) of the fully coupled configuration of CESM2.
+We compared our simulations to the CESM2 Large Ensemble (CESM2-LE) \cite{rodgers2021}, which consists of 100 transient simulations (1850-2100) of the fully coupled configuration of CESM2.
 The simulations follow historical forcing through 2014 and then switch to the SSP3-7.0 emissions scenario.
@@
 We opted to force our simulations with SSTs from ERA5, the fifth generation European Center for Medium Range Weather Forecasts atmospheric reanalysis product \cite{hersbach2020}.
-These data are available at 0.25$^{\circ}$ resolution, which we regridded to the CESM2 nominal 1$^{\circ}$ resolution.
+ERA5 itself used prescribed SSTs which are a blend of HadISST and OSTIA and these data are available at 0.25$^{\circ}$ resolution, which we regridded to the CESM2 nominal 1$^{\circ}$ resolution.
 We likewise utilized the wind fields from ERA5, in order to nudge our simulations to the 2020 circulation (described further in Section \ref{sect:exp}).
@@
 We ran multiple ensembles of simulations using CESM2, each consisting of 30 members, with varied initial conditions. For each simulation, winds were nudged to the ERA5 reanalysis for the year 2020. 
-Nudging occurred four times per day in each grid cell outside the box bounded by 28$^{\circ}$N, 47$^{\circ}$N, 130$^{\circ}$W, 96$^{\circ}$W (Figure \ref{fig:fig1}a).
+Nudging occurred four times per day in each grid cell outside the box bounded by 28$^{\circ}$N, 47$^{\circ}$N, 130$^{\circ}$W, 96$^{\circ}$W (red in Figure \ref{fig:fig1}a).
 This serves to constrain the large-scale circulation, imposing meteorological drought conditions for every ensemble member.
-Inside the bounding box, the atmosphere evolves freely, allowing for local land-atmosphere coupling, which is necessary for studying the effects of initial conditions.
-Sea surface temperatures (SSTs) were imposed from an ERA5 reanalysis product, and all other forcing (e.g.
-CO$_2$, aerosols) followed the standard conventions for transient simulations as in the CESM2-LE.
+Inside the bounding box, the atmosphere evolves freely aside from the constraints imposed by nudging the large scale circulation outside of the box, allowing for local land-atmosphere coupling, which is necessary to study the effects of initial conditions.
+Sea surface temperatures (SSTs) were imposed from ERA5, and all other forcing (e.g. CO$_2$, aerosols) followed the standard conventions for transient simulations as in the CESM2-LE. 
 The simulations were initialized in April, using random initial conditions from the CESM2-LE for all the requisite land and atmosphere states (e.g. temperature, soil moisture, humidity).
@@
 We also simulated this drought during pre-industrial and future conditions to investigate the influence of climate change.
-All of the ensembles were nudged to the 2020 winds, but initial conditions and other climate forcings were drawn from the appropriate time period of the CESM2-LE, effectively porting the 2020 circulation to these two alternative climates.
-SST forcing was anomaly-based, computing anomalies from the CESM2-LE and applying them to the ERA5 reanalysis SSTs used in the 2020 ensemble.
+All of the ensembles were nudged to the 2020 winds, but initial conditions and other climate forcings were drawn from the appropriate time period of the CESM2-LE, which is tantamount to porting the 2020 circulation to these two alternative climates.
+SST forcing for the two alternate periods was anomaly-based, wherein we computed anomalies from the CESM2-LE ensemble mean and applied them to the ERA5 reanalysis SSTs used in the 2020 ensemble.
 Initial conditions were drawn at random from the CESM2-LE from either April 1850 or April 2090, and all other forcing (e.g.
 CO$_2$, aerosols) followed the standard conventions for transient simulations as in the CESM2-LE, which follows the SSP3-7.0 emissions scenario.
-The two key differences between our ensembles and the CESM2-LE are that 1) we impose the 2020 drought circulation via nudging and 2) we forced the simulations with observed SSTs (in lieu of dynamically coupled SSTs).
+The two key differences between our ensembles and the CESM2-LE are that 1) we impose the 2020 drought circulation via nudging and 2) we forced the simulations with SSTs (in lieu of dynamically coupled SSTs).
 
@@
 The CESM2-LE slightly over-predicts precipitation mean and variance in the SWUS, compared to the GPCP reanalysis product across 1981-2020 (Figure \ref{fig:fig1}b).
-Furthermore, CESM2 seems to underestimate the relative contribution of summer rain to annual precipitation and does not accurately match the observed monsoon timing.
+Furthermore, CESM2 underestimates the relative contribution of summer rain to annual precipitation and does not accurately match the observed monsoon timing.
 Despite this, the regional monsoon period (JAS) precipitation statistics match very closely between the CESM2-LE and the GPCP reanalysis product (Supp Figure 3).
 The 2020 SWUS drought was very dry (JAS precip = 71mm, GPCP), which sits at the the 1.8th percentile of the CESM2 large ensemble, primarily driven by extremely dry August and September months.
-Despite the extreme nature of the 2020 monsoon failure, the CCE is able to reproduce the observed meteorological drought.
+Despite the extreme nature of the 2020 monsoon failure, the CCE is able to reproduce the observed meteorological drought, demonstrating an important role for the circulation pattern that occurred in producing it.
 
@@
 \includegraphics[width=35pc]{../figs/main/inits.pdf}
-\caption{The influence of soil water initialization on various summer (JAS) drought indicators in the CCE, shown in the context of the variance across the CESM2-LE.
+\caption{The influence of soil water initialization on various summer (JAS) drought indicators in the 2020 CCE, shown in the context of the variance across the CESM2-LE.
 In each case, the 10cm soil water content at initialization is plotted (red dots) against the summer drought indicator, with linear regression where appropriate (red lines).
@@
 Of the various drought indicators, ET was most closely correlated with soil water initialization (Figure \ref{fig:inits}c), but net biome production experienced the largest initialization effect (based on the number of contours crossed by the regression line, Figure \ref{fig:inits}e), where drier initial soils significantly degraded the carbon sink.
-Monsoon period precipitation itself was not correlated with soil water initialization, indicating that the meteorological drought severity does not appear sensitive to initial conditions, consistent with previous findings \cite{jiang2025}.
+Monsoon period precipitation itself was not correlated with soil water initialization, indicating that the meteorological drought severity is not sensitive to initial conditions in CESM, consistent with previous findings \cite{jiang2025}.
 One more tk sentence about how CCE is pretty nice for testing initial conditions.
@@
 \caption{Summer evapotranspiration vs.
-precipitation in the CCE (a) and a dry subset of the CESM2-LE (d), alongside histograms of ET (b,e) and precipitation (c,f) across three time periods.
+precipitation in the CCEs (a) and a dry subset of the CESM2-LE (d), alongside histograms of ET (b,e) and precipitation (c,f) across three time periods.
 The CESM2-LE is subset to the 30 driest ensemble-years for each period from 1000 total ensemble-years (100 members, 10 years around the given year) based on JAS 10cm soil water content.
@@
 
-We likewise tested the CCE across three climate scenarios: pre-industrial, present-day, and late century in an SSP3-7.0 warming scenario.
-The CCE shows a climate change influence on the regional water cycle, where both ET and precipitation are reduced moving from the pre-industrial period to present-day, and again moving from present-day to the end of this century (Figure \ref{fig:scatter}a-c).
-CCE monsoon precipitation was reduced by 4mm (tk, precise this) on average between the pre-industrial and present-day simulations, with another 9mm reduction expected with SSP3-7.0 by the end of the century.
-Given that the 2020 monsoon failure represented a 64.4mm shortfall relative to climatological precipitation (Supp Figure 3), the CCE scenario effect of -4mm is small relative to the overall drought signal.
-As such, the meteorological drought seems primarily attributable to unforced climate variability \cite{seager2022}.
+We ran CCEs across three climate scenarios: pre-industrial, present-day, and late century in an SSP3-7.0 warming scenario.
+The CCEs show a climate change influence on the regional water cycle, where both ET and precipitation are reduced moving from the pre-industrial period to present-day, and again moving from present-day to the end of this century (Figure \ref{fig:scatter}a-c).
+CCE monsoon precipitation was reduced by 4mm (tk, precise this) on average between the pre-industrial and present-day simulations, with another 9mm reduction simulated with SSP3-7.0 by the end of the century.
+Given that the 2020 monsoon failure represented a 64mm shortfall relative to climatological precipitation (Supp Figure 3), the CCE historical climate change  effect of -4mm is small relative to the overall drought signal.
+We infer, therefore, that the meteorological drought is primarily attributable to unforced climate variability, which is consistent with prior assessments \cite{seager2022}.
 As such, climate change may have exacerbated the monsoon failure, but only slightly, and could potentially increase the severity of future precipitation shortfalls by another 9mm or roughly 12\%. 
-However, the CCE experiment cannot capture any coupled circulation feedbacks that could exacerbate or mitigate the radiative response. 
+However, we note that the CCE experiment cannot capture any coupled circulation feedbacks that could exacerbate or mitigate the radiative response. 
 A tk sentence here providing a literature reference to forced circulation changes.
 
-The CESM2-LE, which does include circulation feedbacks, likewise shows a drying response across the most drought years, but with significantly more noise (Figure \ref{fig:scatter}d-f).
-Because the CCE constrains the large-scale circulation, there is much less variance within each ensemble. 
+The CESM2-LE, which does include circulation feedbacks, likewise shows a drying response across the driest years, but with significantly more noise (Figure \ref{fig:scatter}d-f).
+Because the CCE constrains the large-scale circulation, there is much less variance within each ensemble than in CESM2-LE. 
 In the CESM2-LE, the large-scale circulation evolves freely, yielding a unique circulation pattern for every ensemble member.
 The response of ET to forcing in the CESM2-LE is consistent between PI to present and present to future, both showing a downward trend. 
-Precipitation seems to decline slightly from pre-industrial to present-day, but appears equivalent when comparing present-day and future. 
+Precipitation declines slightly from pre-industrial to present-day, but minimal decline occurs between present-day and the future. 
 These results correspond to a subset of dry years in the CESM2-LE, where we are analyzing the 30 driest monsoons in each time period (based on 10cm soil water), because our interest is specifically in forced changes to extreme droughts.
 Importantly, the forced responses at the extreme quantiles are distinct from those at moderate or wet quantiles (Supp Figure 4), highlighting the importance of investigating extremes separately from mean changes.
-For example, the overall CESM2-LE response to climate change in the SWUS is higher evapotranspiration, but it occurs with a prominent dipole of dry years trending drier, and moderate and wet years trending towards more evaporation.
+For example, the overall CESM2-LE response to climate change in the SWUS is higher evapotranspiration, but it occurs with a prominent dipole of dry years trending towards even less ET, and moderate and wet years trending towards more ET.
 (This dipole seems to be unique with CESM2, but I don't really want to introduce the other datasets at this point, unless we think it's especially important)
@@
 There is a strong relationship between summer soil water and ET in both the CCE and the CESM2-LE (Figure \ref{fig:swcontours}).
-The CCE simulations vary only by the initial conditions, whereas the CESM2-LE data are sampled across various initial conditions and circulations patterns (with strong covariance). 
-Despite this, the two ensembles show comparable ET$\sim$SW relationships, indicating that the ET flux is largely determined by the soil moisture state alone, and relatively agnostic to the specific circulation pattern.
+The CCE simulations vary only by the initial conditions, whereas the CESM2-LE data are sampled across various initial conditions and circulation patterns (with strong covariance). 
+Despite this, the two ensembles show comparable ET$\sim$SW relationships, indicating that the ET flux is largely determined by the soil moisture state, and is relatively agnostic to the specific circulation pattern.
 This is in contrast to the relationship between precipitation and ET, where the slopes of the two relationships look quite different (Figure \ref{fig:pcontours}).
@@
 The large discrepancy in the P$\sim$ET relationship between the CCE and the CESM2-LE seems to indicate that the circulation pattern has a stronger influence on precipitation than does the local water availability.
-As such, a CCE framework allows us to explore how the model would behave absent circulation feedbacks, and in turn diagnose potential circulation feedback results when comparing a CCE to a free-running large ensemble.
+As such, the CCE framework allows us to explore how the model responds to climate change absent circulation feedbacks, and in turn, diagnose potential circulation feedback impacts when comparing the CCE to a free-running large ensemble.
 
 In this case, we ported the 2020 drought circulation to PI and future climates, which imposes the climate change radiative forcings, without any coupled changes to the circulation pattern.
-In many ways the CCE behaves akin to the CESM2-LE, however the CCE appears much drier relative the the CESM2-LE in 2090 than it does in 1850 or 2020 (e.g.
-CCE centroid well outside the CESM2-LE contours in Figure \ref{fig:swcontours}).
-Whereas the consequences of the 2020 drought circulation are within the envelope of the CESM2-LE in the PI and 2020 experiments, porting the circulation to 2090 results in very high values of moisture flux divergence and extremely low surface humidity that exceed the most extreme droughts of the CESM2-LE (Supp Figure 5).
-This would seem to indicate that coupled circulation feedbacks may partially mitigate extreme divergence events in CESM2, under late-century SSP3-7.0 conditions.
+In many ways the CCE behaves akin to the CESM2-LE, however the CCE is much drier relative to the CESM2-LE in 2090 than it is in 1850 or 2020 (e.g., the CCE centroid is well outside the CESM2-LE contours in Figure \ref{fig:swcontours}).
+Whereas the consequences of the 2020 drought circulation are within the envelope of the CESM2-LE in the PI and 2020 experiments, porting the circulation to 2090 results in very high values of moisture flux divergence and extremely low surface humidity, humidity levels that are significantly lower than that occur during the most extreme droughts of the CESM2-LE (Supp Figure 5).
+This result suggests that either the probability of having a circulation pattern that induces a drought as severe as 2020 is reduced or that, when such circulation patterns occur, coupled circulation feedbacks may partially mitigate extreme divergence events in CESM2, under late-century SSP3-7.0 conditions.
 
@@
 \caption{Summer evapotranspiration vs.
-10cm soil water content from the CCE (dots) and the CESM2-LE (scatter density contours) across the three time periods.
-The relationship in the CCE closely follows the CESM2-LE pattern.
-The CCE exists at the dry extreme of the CESM2-LE, or even beyond, as in the 2090 scenario.
+10cm soil water content from the CCEs (dots) and the CESM2-LE (scatter density contours, deeper reds indicate more density) across the three time periods.
+The relationship in the CCEs closely follows the CESM2-LE pattern.
+The CCEs exists at the dry extreme of the CESM2-LE, or even beyond, as in the 2090 scenario.
 }
@@
 
-This study highlights the utility of a constrained circulation ensemble experiment for investigating climate extremes. 
-Our CCE is able to reproduce the 2020 SWUS drought, despite known biases in the fully coupled CESM2.
-This experiment demonstrates that antecedent soil moisture exacerbates drought severity, but not the precipitation amounts.
-We likewise found only a small influence of climate change on precipitation volume during the 2020 drought in the CCE.
-It showed a larger effect by the end of the 20th century, but subject to uncertainty due to potential circulation feedbacks.
+This study highlights the utility of constrained circulation ensemble experiments to investigate climate extremes. 
+Our 2020 CCE is able to reproduce the 2020 SWUS drought, despite known biases in the fully coupled CESM2.
+This experiment demonstrates that antecedent soil moisture exacerbates drought severity, but not the precipitation deficits.
+We likewise found a small, but robust, influence of the thermodynamic impacts on climate change on precipitation amount during the 2020 drought in the CCE.
+That is, despite constraining circulation to be the same in each case, we found that the impacts of that circulation pattern on precipitation was to enhance the associated precipitation deficits.
+An even larger negative effect on precipitation is found by the end of the 21st century.
+Whether such an impact would actually manifest is unclear given the distinction between the behavior of the CCE and the free running CESM2-LE by the end of the 21st century which raises the possibility that forced circulation change and/or circulation feedbacks may limit the occurrence of such extreme drought inducing circulation patterns .
 
@@
 As such, whereas there may have been a small contribution of climate change to the historic precipitation shortfall of 2020 ($\sim$6$\%$ in the CCE), it is not clear that future climate change would further exacerbate monsoon failure in the SWUS.
-While the 1850 and 2020 ensembles looked comparable, in 2090 the CCE is drier than almost every CESM2-LE simulation with regard to 2m vapor pressure and 10cm soil moisture.
+While the 1850 and 2020 ensembles looked comparable, in 2090 the CCE is outside of the range found in the CESM2-LE simulation with regard to 2m vapor pressure and 10cm soil moisture.
 This may indicate a propensity for CESM2 to mitigate extreme moisture flux divergence via circulation feedbacks, but would need to be investigated more specifically.
@@
 \section*{Open Research Section}
-This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
-Data cannot be listed as ''Available from authors'' or stored solely in supporting information.
-Citations to archived data should be included in your reference list.
-Wiley will publish it as a separate section on the paper’s page.
-Examples and complete information are here:
-https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
+We will upload the CCE data to a DOI-minted repository, after the first manuscript review. 
+For the time being, we have uploaded our post-processed data to a temporary location online (\url{https://webext.cgd.ucar.edu/FHIST_BGC/swus.cce.2020/}).
+Data from the GPCP and CESM2-LE are openly available online, but we have likewise uploaded our post-processed data for the four corners region to that website, for ease of review.
+Scripts to set up our simulations, post-process the data, and execute our analyses are online at \url{https://github.com/djk2120/mapp_droughts}.
 

### 2025-11-21 ms/cce.tex
overleaf

@@
 Drought in the SWUS and its intersection with climate change, is challenging to understand from observations alone, and has been frequently investigated with the aid of global or regional climate models.
-While models tend to agree on trends in temperature, soil moisture, and runoff, precipitation is more difficult to predict across this domain, due to the complex orography and the dynamic nature of the North American Monsoon \cite{cook2020,pascale2017}.
+Models tend to agree on trends in temperature, with associated reductions in soil moisture.
+The forced responses of precipitation and runoff are more uncertain.
+Precipitation is difficult to predict across this domain, due to complex orography and the dynamic nature of the North American Monsoon \cite{cook2020,pascale2017}.
 There is also evidence that models may systematically underestimate the drying effects of climate change on near-surface atmospheric vapor pressure in semi-arid regions \cite{simpson2024}.
 As such, it can be difficult to make strong inference regarding precipitation variations or trends in the region, using large-scale coupled climate models.
-The general consensus seems to be that climate change has indeed exacerbated drought in the SWUS in a general sense, but that the precipitation shortfalls of the extreme 2020 drought themselves are primarily attributable to random (natural) atmospheric variations, with potential contributions from a shift towards a sustained negative phase of the Pacific Decadal Oscillation \cite{mankin2021,seager2022, jiang2025}.
+The general consensus seems to be that climate change has indeed exacerbated SWUS drought in a general sense, but that the precipitation shortfalls of the extreme 2020 drought themselves are primarily attributable to random (natural) atmospheric variations, with potential contributions from a shift towards a sustained negative phase of the Pacific Decadal Oscillation \cite{mankin2021,seager2022, jiang2025}.
 
@@
 Instead of one instance of the year 2020, there are 100, such that we can generate large samples of SWUS summers, in order to better analyze drought dynamics in the context of natural variability.
-However, the statistical basis can still become small when focusing explicitly on extreme quantiles, or when controlling for very specific circulation criteria. 
+However, the statistical basis can still become small when focusing explicitly on extreme quantiles, or when filtering for very specific circulation criteria. 
 Attribution of the 2020 SWUS drought is challenging, partly because of our imperfect models, but also because we are trying to diagnose changes not in the mean behavior, but in the severity of rare events.
@@
 
-We opted to force our simulations with SSTs from ERA5, the fifth generation European Center for Medium Range Weather Forecasts atmospheric reanalysis product \cite{hersbach2020}.
-ERA5 itself used prescribed SSTs which are a blend of HadISST and OSTIA and these data are available at 0.25$^{\circ}$ resolution, which we regridded to the CESM2 nominal 1$^{\circ}$ resolution.
-We likewise utilized the wind fields from ERA5, in order to nudge our simulations to the 2020 circulation (described further in Section \ref{sect:exp}).
+We opted to force our simulations with NOAA Extended Reconstructed Sea Surface Temperature V5 \cite{huang2017}.
+These are available at 2$^{\circ}$ resolution, which we regridded to the CESM2 nominal 1$^{\circ}$ resolution.
+We utilized the wind fields from ERA5, the fifth generation European Center for Medium Range Weather Forecasts atmospheric reanalysis product \cite{hersbach2020}, in order to nudge our simulations to the 2020 circulation (described further in Section \ref{sect:exp}).
 All other forcings (e.g. aerosols, CO$_2$, land use) follow the protocol of the CESM2-LE \cite{rodgers2021}.
@@
 \begin{figure}
-\includegraphics[width=35pc]{../figs/main/contours_ET_SW.pdf}
+\includegraphics[width=35pc]{../figs/main/fig4}
 \caption{Summer evapotranspiration vs.
@@
 }
-\label{fig:swcontours}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=35pc]{../figs/main/contours_PREC_ET.pdf}
-\caption{Summer precipitation vs.
-evapotranspiration from the CCE (dots) and the CESM2-LE (scatter density contours) across the three time periods.
-The location of the CCE data comports with the CESM2-LE pattern, but the slopes do not appear to match.
-N.b.
-will combine this with Figure 5.}
-\label{fig:pcontours}
+\label{fig:contours}
 \end{figure}

### 2025-11-24 postp/avg_lens2_foco_daily.sh
more postp stuff

@@ -1,0 +1,7 @@
+indir='/glade/derecho/scratch/djk2120/postp/lens2/daily/'
+outdir='/glade/derecho/scratch/djk2120/postp/lens2/daily_avg/'
+
+while read year; do
+    job='avg_foco_'$year'.job'
+    echo $year
+done<lens2.years

### 2025-11-24 postp/cce_subset_foco_mon.py

@@ -1,0 +1,40 @@
+import xarray as xr
+from utils import get_files
+import sys
+
+# inputs
+year = int(sys.argv[1])
+mem = int(sys.argv[2])
+exp = str(year) + '_era5_i04'
+
+# subsetting paradigm
+a = slice(30, 45)
+o = slice(240, 265)
+dvs = ['RAIN', 'SNOW', 'QVEGT', 'QVEGE', 'QSOIL', 'QRUNOFF',
+       'RH2M', 'TSA', 'SOILWATER_10CM']
+advs = ['TMQ', 'PBLH']
+
+# land data
+files = get_files(exp)
+mstr = files[mem][0].split('clm2')[0][-4:-1]
+dsets = [xr.open_dataset(f, decode_timedelta=False)[dvs].sel(lat=a, lon=o)
+         for f in files[mem]]
+ds = xr.concat(dsets, dim='time').compute()
+ds['time'] = xr.date_range(str(year), freq='MS', periods=12)[3:]
+tmp = xr.open_dataset(files[mem][0], decode_timedelta=False)
+ds['landarea'] = tmp.area*tmp.landfrac
+
+# atm data
+afiles = get_files(exp, cmp='/atm/')
+dsets = [xr.open_dataset(f, decode_timedelta=False)[advs].sel(lat=a, lon=o)
+         for f in afiles[mem]]
+ds_atm = xr.concat(dsets, dim='time').compute()
+ds_atm['lat'] = ds.lat
+ds_atm['time'] = ds.time
+for v in ds_atm.data_vars:
+    ds[v] = ds_atm[v]
+
+# write to file
+dout = '/glade/derecho/scratch/djk2120/postp/cce/'
+fout = 'cce.{}.foco.gridded.mon.{}.nc'.format(exp, mstr)
+ds.to_netcdf(dout + fout)

### 2025-11-24 postp/lens2_subset_foco_daily.py

@@ -1,0 +1,62 @@
+import xarray as xr
+import glob
+import sys
+import numpy as np
+
+# inputs
+year = int(sys.argv[1])
+mem = sys.argv[2]
+
+# data vars
+avs = ['TMQ', 'TREFHTMX', 'PBLH', 'TREFHT', 'RHREFHT']
+lvs = ['RAIN', 'SNOW', 'QVEGT', 'QVEGE', 'QSOIL', 'QRUNOFF',
+       'SOILWATER_10CM', 'FSH', 'GPP', 'TLAI']
+dvs = [*avs, *lvs]
+
+# file key
+lens2 = '/glade/campaign/cgd/cesm/CESM2-LE/'
+lnd_dir = lens2 + 'timeseries/lnd/proc/tseries/day_1/'
+atm_dir = lens2 + 'timeseries/atm/proc/tseries/day_1/'
+afile = atm_dir+'{}/*{}*h1*{}*.nc'  # glob pattern
+lfile = lnd_dir+'{}/*{}*h5*{}*.nc'
+f0 = {v: afile for v in avs}
+for v in lvs:
+    f0[v] = lfile
+
+# define subsetting patterns
+a = slice(30, 45)
+o = slice(240, 265)
+
+# read data
+dsout = xr.Dataset()
+for v in dvs:
+    g = f0[v].format(v, mem, year)
+    files = glob.glob(g)
+    if len(files) == 1:
+        ds = xr.open_dataset(files[0])
+        da = ds[v].sel(lat=a, lon=o).compute()
+    else:
+        print('empty glob: '+g)
+        da = np.nan*da
+    da['lat'] = np.round(da.lat.astype(float), 2)  # lnd/atm mismatch
+    dsout[v] = da
+
+# derived variables
+t = dsout.TREFHT-273.15
+rh = dsout.RHREFHT/100
+es = 0.61094*np.exp(17.625*t/(t+234.04))
+note = 'calc from daily TREFHT and RHREFHT'
+dsout['VPD'] = ((1-rh)*es).compute()
+dsout['VPD'].attrs = {'long_name': 'vapor pressure deficit',
+                      'units': 'kPa', 'note': note}
+dsout['VP'] = (rh*es).compute()
+dsout['VP'].attrs = {'long_name': 'vapor pressure',
+                     'units': 'kPa', 'note': note}
+dsout['PREC'] = dsout['RAIN'] + dsout['SNOW']
[…]

### 2025-11-24 postp/subset_cce_mon.sh

@@ -1,0 +1,9 @@
+
+for year in "2020" "1850" "2090"; do
+    for i in {0..29}; do
+        job="cce.foco."$year"."$i".job"
+        sed "s/year/"$year"/g" subset_cce_mon.template > $job
+        sed -i "s/mem/"$i"/g" $job
+	qsub $job
+    done
+done

### 2025-11-24 postp/subset_lens2_daily.sh

@@ -1,0 +1,19 @@
+i=0
+j=0
+while read mem; do
+    while read year; do
+	((i++))
+	if [[ $i == 11 || $i == 1 ]]; then
+	    i=1
+	    ((j++))
+	    num=$(printf %03d $j)
+	    job="cesm2.foco.daily."$num".job"
+	    sed 's/num/'$num'/g' subset_lens2_daily.template > $job
+	fi
+	line="python lens2_subset_foco_daily.py "$year" "$mem
+	echo $line >> $job
+	if [[ $i == 10 ]]; then
+	    qsub $job
+	fi
+    done < lens2.years
+done < lens2.mems

### 2025-11-24 postp/avg_cce_foco_daily.sh
more postp

@@ -1,0 +1,10 @@
+indir='/glade/derecho/scratch/djk2120/postp/cce/daily/'
+outdir='/glade/derecho/scratch/djk2120/postp/cce/daily_avg/'
+
+for year in 2020 1850 2090; do
+    job='avg_foco_'$year'.job'
+    sed 's/year/'$year'/g' avg_cce_foco.template > $job
+    sed -i 's:indir:'$indir':g' $job
+    sed -i 's:outdir:'$outdir':g' $job
+    qsub $job
+done

### 2025-11-24 postp/avg_lens2_foco_daily.sh

@@ -3,5 +3,8 @@ outdir='/glade/derecho/scratch/djk2120/postp/lens2/daily_avg/'
 
 while read year; do
     job='avg_foco_'$year'.job'
-    echo $year
+    sed 's/year/'$year'/g' avg_foco.template > $job
+    sed -i 's:indir:'$indir':g' $job
+    sed -i 's:outdir:'$outdir':g' $job
+    qsub $job
 done<lens2.years
